### 2021-09-09 paper.Rmd
initial setup

---
title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
author: |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.2
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}

bibliography: references.bib
csl: american-journal-of-political-science.csl
output:
  bookdown::pdf_document2:
    toc: no
    keep_tex: true
mainfont: Times New Roman
sansfont: Times New Roman
fontsize: 12pt
link-citations: true
documentclass: article
geometry: margin=1in
always_allow_html: yes

---

```{r, include=FALSE}
# packages
library(tidyverse)
library(broom)
library(scales)
library(modelsummary)
library(gridExtra)
library(jtools)
library(tidyverse)
library(sjmisc)
library(scales)
```

\newpage
# Introduction
<!-- Guiding Question about people, not scholars! -->
How do emotional appeals interact with corrective information?
<!-- context on misinformation and the important role of emotions in spreading it -->
Fueled by the fast dissemination through digital media, the issue of various forms of misinformation has become a major concern both in academia and for public policy [@jerit2020political; @lazer2018science; @lewandowsky2012misinformation; @wardle2017information].
To limit the danger from inaccurate or misleading information, efforts of fact-checking or corrective information more generally are widely applied [@walter2020fact]. A common finding in the maturing literature on corrective information is that these interventions are usually successful in reducing misperceptions, but may not translate to other outcomes such as attitudes, candidate evaluations or blame attribution [@bisgaard2019getting; @nyhan2019taking; @swire2019they; @thorson2016belief].

<!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2020misinformation; @weeks2015emotions]. This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread]. In general, moral-emotional language is rewarded on social media [@brady2017emotion] and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how].

<!-- % emotion in political science -->
Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions. 
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization]. 
Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude]. 
While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].

<!-- % get more concrete linking belief echoes / affective perseverance and specific emotion with possible connection to moralization -->
In his study, Clifford shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes. However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer. Such a piece of information can be classified as misinformation in the sense of going against ``best available evidence in the public sphere''[@flynn2017nature, p.128].
From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.

<!-- % Contribution bringing belief echoes and discrete emotions together -->
This study contributes to the literature of corrective information by linking the concept of belief echoes to the perseverance of specific emotions with related consequences on attitudes. Specifically, it provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims.

<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, hasell2020differential, carnahan2019processing]. This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans]. GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].

<!-- % Final Punchline on broader relevance -->
In the view of many scientists [@roberts2018nobel, planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.

# Theory

## Misinformation and corrective information

<!-- % some definitions from Kuklinski and Flynn to Jerit -->

<!-- % information corrects factual beliefs, but the effects of information on attitudes are mixed -->

Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information.

The effects on related opinions are much less clear.
Some studies find effects, some not.

Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected.

## Emotions and political learning

<!-- % From Vasiloupolou / Wagner -->
<!-- %Early research into emotional reactions either used one simple positive–negative dimension or a two-dimensional valence model (Marcus, 2003; Marcus et al., 2000). More recently, researchers have argued that distinct emotions may characterise individuals’ reactions (e.g. Huddy et al., 2007; Larsen and McGraw, 2011; Petersen, 2010; Smith et al., 2008; Valentino et al., 2008, 2011; Weber, 2012; see also Conover and Feldman, 1986). Appraisal theories of emotions suggest that various distinct emotions arise due to individuals’ cognitive interpretation of the situation they experience (e.g. Roseman, 1991; Smith and Ellsworth, 1985; for a review, see Brader and Marcus, 2013). Neuroscientific theories based on the neural processes that generate emotional responses have also increasingly turned towards a three-dimensional model of emotions that distinguishes between fear, anger and enthusiasm (Brader and Marcus, 2013; Lerner and Keltner, 2001). Research in political psychology building on both appraisal and neural process theories has therefore focused in particular on the causes and consequences of anger, fear1 and enthusiasm as distinct emotions (Carver and Harmon-Jones, 2009; Conover and Feldman, 1986; Huddy et al., 2005, 2007; Lerner and Keltner, 2000, 2001; Petersen, 2010; Smith and Ellsworth, 1985; Valentino et al., 2011). -->
<!-- % -->
<!-- %Applied to political phenomena, distinct emotions cause different types of actions and attitudes: fear leads to caution and openness, and anger to more confident and aggressive responses (Druckman and McDermott, 2008; MacKuen et al., 2010). For instance, whereas fear and anger both tend to increase intolerance and threat perceptions (e.g. Erişen and Kentmen-Cin, 2017), they lead to distinct ways of reacting to terrorist threats (Lerner et al., 2003). Fearful citizens are more wary of the increased risk and turn to isolationism, while angry citizens are less risk-averse and tend to favour pro-active intervention (Huddy et al., 2007). Anxious citizens are also more likely to seek out new information and process this carefully (Brader et al., 2008; Marcus et al., 2000; Merolla and Zechmeister, 2009; Valentino et al., 2008). Finally, enthusiasm increases the willingness to participate in politics, but like anger it also increases less careful forms of information-processing (Brader, 2006; Marcus et al., 2000). -->

Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.

Research has shown that... anger... anxiety... enthusiasms...

A recently growing interest is in disgust.
Connection to morality, relevant for immigration, homeless ...

A study has shown that moral shock works to moralize attitudes towards GM food.
However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

## Hypotheses

From findings on moral conviction and how emotional frames induce it [@clifford2019emotional; @wisneski2017moralization], I expect that a persuasive message evoking negative emotions, notably disgust and anger, can increase perceptions of the respective issue as being one of moral conviction. In order to count as a persuasive message, I expect the message to affect both factual beliefs and policy opinions in its direction.

* H1 A persuasive, emotionally framed message evoking negative emotions of disgust and anger makes individuals more likely to favor policy restrictions on the respective issue and to perceive the issue as one of moral conviction.

<!-- % Moderation -->

<!-- % possible theoretical contribution by categorizing different emotion effects -->
<!-- %To categorize effects of emotions within the context mis- and corrective information, there are two general ways in which emotions may affect its effectiveness. In both cases, one can further differentiate between cases where the emotion is invoked by, or at least closely tied to, the object that is part of the information (integral emotion), and cases where the emotion is induced irrespective of the piece of information (incidental emotion, often also referred to as mood). Emotion as moderator. emotions affect information processing so also for misinformation and corrective information. But anxiety, anger and disgust word in different directions... emotion within fake news, see Martel. Emotion as outcome -->

Concerning the impact of corrective information, I expect that it will be effective in reducing factual misperceptions, irrespective of the presence of evoked emotions. This is in line with previous research showing no moderation effect on corrections due to emotional framing  [@ecker2011terrorists].

* H2a Corrective information is effective in eliminating the misinformation effect on factual beliefs.

<!-- % Belief echoes and affective perseverance -->

However, similar to previous results on affective perseverance [@sherman2002affective] and belief echoes [@thorson2016belief], I expect the effects of emotionally framed misinformation on policy opinion, moral conviction, and emotions to persist even in the face of effective factual correction.

* H2b The emotionally framed misinformation leaves an impact on policy opinions, moral conviction, and emotions.

Together, hypotheses 2a and 2b represent the existence of a ``belief echo'', i.e. the continued effect of a piece of misinformation on attitudinal or emotional outcomes, despite successful factual correction. If only H2a is supported, the correction is altogether successful.

<!-- % Counter-framing -->

Corrective information such as fact-checking is usually deliberately presented in a rather neutral way. In the case of strongly emotional framed misinformation, such a neutral correction may thus be successful in correcting factual beliefs, but not be able to counter the emotional impact of the misinformation. If this is the case, a more effective strategy may be to counter the emotional misinformation not with a neutral correction but with a correction embedded in an emotional counter-frame. If equally persuasive, such counterframes have been shown to negate the effect of an original frame [@chong2013counterframing]. I expect a similar effect to hold in response to emotionally framed misinformation.

* H3 A correction embedded in a persuasive emotional counter-frame eliminates the misinformation effect on policy opinions, moral conviction, and emotions.

<!-- %Visual cues relate to \citep{hameleers2020picture} and \citep{powell2021selection} -->

## GMO - a case for emotional reactions and misinformation

# Data & Methods

# Results

## Descriptives

```{r, warning=FALSE, echo=FALSE}
# load data
source("data-preparation.R")
```

```{r, warning=FALSE, echo=FALSE}
# only control group

descriptives_control <- data %>% 
  filter(exp_group == "control") %>% 
  select("Anger" = "anger",
         "Anxiety" = "anxiety",
         "Disgust" = "disgust",
         "Sadness" = "sadness",
         "Belief GMO safety" = "knowledge_safe", 
         "Policy restrictions" = "policy_opinion",
         "Moral conviction" = "moral_conviction")

datasummary_skim(descriptives_control)

```

## Moralization

```{r, echo = FALSE}
# linear regression models with 0-1 rescaled DV
lm_knowledge_safe <- lm(rescale(knowledge_safe) ~ exp_group, data)
lm_policy_opinion <- lm(rescale(policy_opinion) ~ exp_group, data)
lm_moral_conviction <- lm(rescale(moral_conviction) ~ exp_group, data)

lm_emotion_disgust <- lm(rescale(disgust) ~ exp_group, data)
lm_emotion_anger <- lm(rescale(anger) ~ exp_group, data)
lm_emotion_anxiety <- lm(rescale(anxiety) ~ exp_group, data)
lm_emotion_sadness <- lm(rescale(sadness) ~ exp_group, data)

# define functions
# function to get coefficients for plotting
get_pdata <- function(model, positions = c(), label = NA) {
  df <- data.frame(summary(model)$coefficients[positions,],
                   group = str_remove(rownames(summary(model)$coefficients[positions,]), "exp_group"),
                   outcome = label)
  df$group <- fct_inorder(df$group)
  df$outcome <- fct_inorder(df$outcome)
  names(df) <- c("estimate", "se", "t", "p", "group", "outcome")
  df$lower <- df$estimate - 1.96*df$se
  df$upper <- df$estimate + 1.96*df$se
  row.names(df) <- NULL
  return(df)
}

```

```{r, echo = FALSE}
# H1: misinformation effect
pdata_h1_belief_policy <- rbind(get_pdata(lm_moral_conviction, c(4, 7), "Moral conviction"),
                                get_pdata(lm_policy_opinion, c(4, 7), "Policy restrictions"),
                                get_pdata(lm_knowledge_safe, c(4, 7), "Belief GMO safety"))

pdata_h1_emotions <- rbind(get_pdata(lm_emotion_sadness, c(4, 7), "Sadness"),
                           get_pdata(lm_emotion_disgust, c(4, 7), "Disgust"),
                           get_pdata(lm_emotion_anxiety, c(4, 7), "Anxiety"),
                           get_pdata(lm_emotion_anger, c(4, 7), "Anger")
                           )

pdata_h1 <- rbind(cbind(pdata_h1_belief_policy, type = "Effect on belief and policy opinion"),
                  cbind(pdata_h1_emotions, type = "Effect on emotions"))

pdata_h1 <- pdata_h1 %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), 
                          "Low emotional misinformation", "High emotional misinformation"))

# actual plotting
ggplot(pdata_h1, aes(x = outcome, y = estimate, color = "1")) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, linetype = emotion),
                size = 0.5,
                width = 0.25,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(fct_rev(emotion) ~ fct_rev(type), scales = "free") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0, end = 0.5, direction = 1) +
  scale_linetype_manual(values=c("solid", "twodash")) +
  scale_shape_manual(values = c(17, 16)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm'),
        strip.text.x = element_text(size = 8)
  ) +
  guides(color = "none",
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

#sjPlot::save_plot("./figures/p_h1.png", width = 24, height = 16)

```

```{r, echo = FALSE}

# neutral correction for both low and high misinformation

pdata_neutral_correction_belief_opinion <- rbind(get_pdata(lm_moral_conviction, c(4, 5, 7, 8), "Moral conviction"),
                                           get_pdata(lm_policy_opinion, c(4, 5, 7, 8), "Policy restrictions"),
                                           get_pdata(lm_knowledge_safe, c(4, 5, 7, 8), "Belief GMO safety"))

pdata_neutral_correction_emotions <- rbind(
  get_pdata(lm_emotion_sadness, c(4, 5, 7, 8), "Sadness"),
  get_pdata(lm_emotion_disgust, c(4, 5, 7, 8), "Disgust"),
  get_pdata(lm_emotion_anxiety, c(4, 5, 7, 8), "Anxiety"),
  get_pdata(lm_emotion_anger, c(4, 5, 7, 8), "Anger")
  )

pdata_neutral_correction <- rbind(cbind(pdata_neutral_correction_belief_opinion, type = "Effect on belief and policy opinion"),
                            cbind(pdata_neutral_correction_emotions, type = "Effect on emotions"))

pdata_neutral_correction <- pdata_neutral_correction %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
         correction = ifelse(str_detect(group, "correction"), "Neutral correction", "No correction"))

# actual plotting
ggplot(pdata_neutral_correction, aes(x = outcome, y = estimate, color = correction, shape = correction)) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, linetype = emotion),
                size = 0.5,
                width = 0.5,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(fct_rev(emotion) ~ fct_rev(type), scales = "free") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0, end = 0.5, direction = -1) +
  scale_linetype_manual(values=c("solid", "twodash")) +
  scale_shape_manual(values = c(17, 16)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm'),
        strip.text.x = element_text(size = 8)
    ) +
  guides(color = guide_legend(reverse = TRUE),
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

#sjPlot::save_plot("./figures/p_neutral_correction.png", width = 24, height = 16)

# counter-frame correction for both low and high misinformation
pdata_counter_frame_belief_opinion <- rbind(
  get_pdata(lm_moral_conviction, c(4, 6, 7, 9), "Moral conviction"),
  get_pdata(lm_policy_opinion, c(4, 6, 7, 9), "Policy restrictions"),
  get_pdata(lm_knowledge_safe, c(4, 6, 7, 9), "Belief GMO safety")
  )

pdata_counter_frame_emotions <- rbind(
  get_pdata(lm_emotion_sadness, c(4, 6, 7, 9), "Sadness"),
  get_pdata(lm_emotion_disgust, c(4, 6, 7, 9), "Disgust"),
  get_pdata(lm_emotion_anxiety, c(4, 6, 7, 9), "Anxiety"),
  get_pdata(lm_emotion_anger, c(4, 6, 7, 9), "Anger")
  )

pdata_counter_frame <- rbind(cbind(pdata_counter_frame_belief_opinion, type = "Effect on belief and policy opinion"),
                                  cbind(pdata_counter_frame_emotions, type = "Effect on Emotions"))

pdata_counter_frame <- pdata_counter_frame %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
         correction = ifelse(str_detect(group, "counter"), "Counter-frame correction", "No correction"))

# actual plot
ggplot(pdata_counter_frame, aes(x = outcome, y = estimate, color = correction, shape = correction)) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, linetype = emotion),
                size = 0.5,
                width = 0.5,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(fct_rev(emotion) ~ fct_rev(type), scales = "free") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0, end = 0.75, direction = -1) +
  scale_linetype_manual(values=c("solid", "twodash")) +
  scale_shape_manual(values = c(15, 16)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm'),
        strip.text.x = element_text(size = 8)
  ) +
  guides(color = guide_legend(reverse = TRUE),
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

#sjPlot::save_plot("./figures/p_counter_frame.png", width = 24, height = 16)

# Appendix
# info only

pdata_info_effect <- rbind(
  get_pdata(lm_moral_conviction, c(2, 3), "Moral conviction"),
  get_pdata(lm_policy_opinion, c(2, 3), "Policy restrictions"),
  get_pdata(lm_knowledge_safe, c(2, 3), "Belief GMO safety"))

pdata_info_effect_emotions <- rbind(
  get_pdata(lm_emotion_sadness, c(2, 3), "Sadness"),
  get_pdata(lm_emotion_disgust, c(2, 3), "Disgust"),
  get_pdata(lm_emotion_anxiety, c(2, 3), "Anxiety"),
  get_pdata(lm_emotion_anger, c(2, 3), "Anger")
  )

pdata_info_effect <- rbind(cbind(pdata_info_effect, type = "Effect on belief and policy opinion"),
                           cbind(pdata_info_effect_emotions, type = "Effect on Emotions"))

pdata_info_effect <- pdata_info_effect %>% 
  mutate(group = fct_recode(group, 
                            "Neutral correction" = "neutral-correction",
                            "Counterframe" = "counter-frame"))

ggplot(pdata_info_effect, aes(x = outcome, y = estimate, color = fct_rev(group), shape = fct_rev(group))) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                size = 0.5,
                width = 0.5,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(~fct_rev(type), scales = "free_y") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0.25, end = 0.75, direction = -1) +
  scale_linetype_manual(values=c("dashed", "solid")) +
  scale_shape_manual(values = c(15, 17)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm')) +
  guides(color = guide_legend(reverse = TRUE),
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

#sjPlot::save_plot("./figures/p_info_effect.png", width = 24, height = 9)

```

# Conclusion

\newpage

# References

\linespread{1}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

::: {#refs}

:::

### 2021-09-09 paper.tex
initial setup

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
  \setsansfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Emotional framing and the effectiveness of corrective information},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=Blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Emotional framing and the effectiveness of corrective information\footnote{I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim's Graduate School of Economics and Social Sciences.}}
\author{Pirmin Stöckle\footnote{Corresponding address: \href{mailto:pirmin.stoeckle@gess.uni-mannheim.de}{\nolinkurl{pirmin.stoeckle@gess.uni-mannheim.de}}}\\
GESS and SFB 884, University of Mannheim}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
\setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
\end{abstract}

\setstretch{1.2}
\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

How do emotional appeals interact with corrective information?
Fueled by the fast dissemination through digital media, the issue of various forms of misinformation has become a major concern both in academia and for public policy (Jerit and Zhao \protect\hyperlink{ref-jerit2020political}{2020}; Lazer et al. \protect\hyperlink{ref-lazer2018science}{2018}; Lewandowsky et al. \protect\hyperlink{ref-lewandowsky2012misinformation}{2012}; Wardle and Derakhshan \protect\hyperlink{ref-wardle2017information}{2017}).
To limit the danger from inaccurate or misleading information, efforts of fact-checking or corrective information more generally are widely applied (Walter et al. \protect\hyperlink{ref-walter2020fact}{2020}). A common finding in the maturing literature on corrective information is that these interventions are usually successful in reducing misperceptions, but may not translate to other outcomes such as attitudes, candidate evaluations or blame attribution (Bisgaard \protect\hyperlink{ref-bisgaard2019getting}{2019}; Nyhan et al. \protect\hyperlink{ref-nyhan2019taking}{2019}; Swire-Thompson et al. \protect\hyperlink{ref-swire2019they}{2019}; Thorson \protect\hyperlink{ref-thorson2016belief}{2016}).

While this has been linked to affect (Thorson \protect\hyperlink{ref-thorson2016belief}{2016}), the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include (Ecker, Lewandowsky, and Apai \protect\hyperlink{ref-ecker2011terrorists}{2011}; Martel, Pennycook, and Rand \protect\hyperlink{ref-martel2020reliance}{2020}; Rosenzweig et al. \protect\hyperlink{ref-rosenzweig2020misinformation}{2020}; Weeks \protect\hyperlink{ref-weeks2015emotions}{2015}). This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust (Vosoughi, Roy, and Aral \protect\hyperlink{ref-vosoughi2018spread}{2018}). In general, moral-emotional language is rewarded on social media (Brady et al. \protect\hyperlink{ref-brady2017emotion}{2017}) and populist parties use more negative emotional appeals in their communication than non-populist parties (Widmann \protect\hyperlink{ref-widmann2021how}{2021}).

Emotions certainly play an important role in politics (Redlawsk \protect\hyperlink{ref-redlawsk2006feeling}{2006}), and approaches such as cognitive appraisal theory (Moors et al. \protect\hyperlink{ref-moors2013appraisal}{2013}) or affective intelligence theory (Marcus, Neuman, and MacKuen \protect\hyperlink{ref-marcus2000affective}{2000}; Marcus et al. \protect\hyperlink{ref-marcus2019applying}{2019}) provide frameworks for empirical expectaitoins about specific emotions.
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, Clifford (\protect\hyperlink{ref-clifford2019emotional}{2019}) finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction (Skitka et al. \protect\hyperlink{ref-skitka2020psychology}{2020}; Wisneski and Skitka \protect\hyperlink{ref-wisneski2017moralization}{2017}).
Moral conviction, i.e.~the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength (Skitka, Wisneski, and Brandt \protect\hyperlink{ref-skitka2018attitude}{2018}).
While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues (Ryan \protect\hyperlink{ref-ryan2014reconsidering}{2014}) with distinct behavioral consequences, such as an absolute unwillingness to compromise (Ryan \protect\hyperlink{ref-ryan2017no}{2017}) or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences (Ryan \protect\hyperlink{ref-ryan2019actions}{2019}).

In his study, Clifford shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes. However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer. Such a piece of information can be classified as misinformation in the sense of going against ``best available evidence in the public sphere''(Flynn, Nyhan, and Reifler \protect\hyperlink{ref-flynn2017nature}{2017}, 128).
From findings of belief echoes (Thorson \protect\hyperlink{ref-thorson2016belief}{2016}) and affective perseverance (Sherman and Kim \protect\hyperlink{ref-sherman2002affective}{2002}) we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.

This study contributes to the literature of corrective information by linking the concept of belief echoes to the perseverance of specific emotions with related consequences on attitudes. Specifically, it provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims.

The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information (Diamond, Bernauer, and Mayer \protect\hyperlink{ref-diamond2020does}{2020}, hasell2020differential, carnahan2019processing). This is especially true within the German population, which is where this study draws its sample from (Gaskell et al. \protect\hyperlink{ref-gaskell2010europeans}{2010}). GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it (Fernbach et al. \protect\hyperlink{ref-fernbach2019extreme}{2019}), and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system (Siegrist and Hartmann \protect\hyperlink{ref-siegrist2020consumer}{2020}).

In the view of many scientists (Roberts \protect\hyperlink{ref-roberts2018nobel}{2018}, planck2019scientists), genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.

\hypertarget{theory}{%
\section{Theory}\label{theory}}

\hypertarget{misinformation-and-corrective-information}{%
\subsection{Misinformation and corrective information}\label{misinformation-and-corrective-information}}

Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information.

The effects on related opinions are much less clear.
Some studies find effects, some not.

Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected.

\hypertarget{emotions-and-political-learning}{%
\subsection{Emotions and political learning}\label{emotions-and-political-learning}}

Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.

Research has shown that\ldots{} anger\ldots{} anxiety\ldots{} enthusiasms\ldots{}

A recently growing interest is in disgust.
Connection to morality, relevant for immigration, homeless \ldots{}

A study has shown that moral shock works to moralize attitudes towards GM food.
However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

\hypertarget{hypotheses}{%
\subsection{Hypotheses}\label{hypotheses}}

From findings on moral conviction and how emotional frames induce it (Clifford \protect\hyperlink{ref-clifford2019emotional}{2019}; Wisneski and Skitka \protect\hyperlink{ref-wisneski2017moralization}{2017}), I expect that a persuasive message evoking negative emotions, notably disgust and anger, can increase perceptions of the respective issue as being one of moral conviction. In order to count as a persuasive message, I expect the message to affect both factual beliefs and policy opinions in its direction.

\begin{itemize}
\tightlist
\item
  H1 A persuasive, emotionally framed message evoking negative emotions of disgust and anger makes individuals more likely to favor policy restrictions on the respective issue and to perceive the issue as one of moral conviction.
\end{itemize}

Concerning the impact of corrective information, I expect that it will be effective in reducing factual misperceptions, irrespective of the presence of evoked emotions. This is in line with previous research showing no moderation effect on corrections due to emotional framing (Ecker, Lewandowsky, and Apai \protect\hyperlink{ref-ecker2011terrorists}{2011}).

\begin{itemize}
\tightlist
\item
  H2a Corrective information is effective in eliminating the misinformation effect on factual beliefs.
\end{itemize}

However, similar to previous results on affective perseverance (Sherman and Kim \protect\hyperlink{ref-sherman2002affective}{2002}) and belief echoes (Thorson \protect\hyperlink{ref-thorson2016belief}{2016}), I expect the effects of emotionally framed misinformation on policy opinion, moral conviction, and emotions to persist even in the face of effective factual correction.

\begin{itemize}
\tightlist
\item
  H2b The emotionally framed misinformation leaves an impact on policy opinions, moral conviction, and emotions.
\end{itemize}

Together, hypotheses 2a and 2b represent the existence of a ``belief echo'', i.e.~the continued effect of a piece of misinformation on attitudinal or emotional outcomes, despite successful factual correction. If only H2a is supported, the correction is altogether successful.

Corrective information such as fact-checking is usually deliberately presented in a rather neutral way. In the case of strongly emotional framed misinformation, such a neutral correction may thus be successful in correcting factual beliefs, but not be able to counter the emotional impact of the misinformation. If this is the case, a more effective strategy may be to counter the emotional misinformation not with a neutral correction but with a correction embedded in an emotional counter-frame. If equally persuasive, such counterframes have been shown to negate the effect of an original frame (Chong and Druckman \protect\hyperlink{ref-chong2013counterframing}{2013}). I expect a similar effect to hold in response to emotionally framed misinformation.

\begin{itemize}
\tightlist
\item
  H3 A correction embedded in a persuasive emotional counter-frame eliminates the misinformation effect on policy opinions, moral conviction, and emotions.
\end{itemize}

\hypertarget{gmo---a-case-for-emotional-reactions-and-misinformation}{%
\subsection{GMO - a case for emotional reactions and misinformation}\label{gmo---a-case-for-emotional-reactions-and-misinformation}}

\hypertarget{data-methods}{%
\section{Data \& Methods}\label{data-methods}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{descriptives}{%
\subsection{Descriptives}\label{descriptives}}

\begin{table}
\centering
\begin{tabular}[t]{lrrrrrrr>{}r}
\toprule
  & Unique (\#) & Missing (\%) & Mean & SD & Min & Median & Max &   \\
\midrule
Anger & 5 & 0 & 1.7 & 1.4 & 0.0 & 2.0 & 4.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_2214ac55bc6.pdf}\\
Anxiety & 5 & 0 & 2.4 & 1.2 & 0.0 & 3.0 & 4.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_22147c85086.pdf}\\
Disgust & 5 & 0 & 1.5 & 1.3 & 0.0 & 1.0 & 4.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_22147726477d.pdf}\\
Sadness & 5 & 0 & 1.7 & 1.3 & 0.0 & 2.0 & 4.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_221427ce1c39.pdf}\\
Belief GMO safety & 8 & 0 & 3.2 & 1.6 & 1.0 & 3.0 & 7.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_2214d7d1cfe.pdf}\\
Policy restrictions & 8 & 1 & 5.2 & 1.6 & 1.0 & 5.0 & 7.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_221475322953.pdf}\\
Moral conviction & 10 & 1 & 4.4 & 2.0 & 0.0 & 4.0 & 8.0 & \includegraphics[width=0.67in, height=0.17in]{C:/Users/pstoeckl/Documents/GitHub_Projects/emotions-corrective-information/paper_files/figure-latex/hist_22145de319c0.pdf}\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{moralization}{%
\subsection{Moralization}\label{moralization}}

\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}

\includegraphics{paper_files/figure-latex/unnamed-chunk-6-1.pdf} \includegraphics{paper_files/figure-latex/unnamed-chunk-6-2.pdf} \includegraphics{paper_files/figure-latex/unnamed-chunk-6-3.pdf}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\linespread{1}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}

\noindent

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-bisgaard2019getting}{}%
Bisgaard, Martin. 2019. ``How Getting the Facts Right Can Fuel Partisan-Motivated Reasoning.'' \emph{American Journal of Political Science} 63(4): 824--839.

\leavevmode\hypertarget{ref-brady2017emotion}{}%
Brady, William J. et al. 2017. ``Emotion shapes the diffusion of moralized content in social networks.'' \emph{Proceedings of the National Academy of Sciences} 114(28): 7313--7318. \url{https://www.pnas.org/content/114/28/7313.short}.

\leavevmode\hypertarget{ref-chong2013counterframing}{}%
Chong, Dennis, and James N. Druckman. 2013. ``Counterframing effects.'' \emph{The Journal of Politics} 75(1): 1--16.

\leavevmode\hypertarget{ref-clifford2019emotional}{}%
Clifford, Scott. 2019. ``How Emotional Frames Moralize and Polarize Political Attitudes.'' \emph{Political Psychology} 40(1): 75--91. \url{https://onlinelibrary.wiley.com/doi/full/10.1111/pops.12507}.

\leavevmode\hypertarget{ref-diamond2020does}{}%
Diamond, Emily, Thomas Bernauer, and Frederick Mayer. 2020. ``Does providing scientific information affect climate change and GMO policy preferences of the mass public? Insights from survey experiments in Germany and the United States.'' \emph{Environmental Politics}: 1--20.

\leavevmode\hypertarget{ref-ecker2011terrorists}{}%
Ecker, Ullrich K. H., Stephan Lewandowsky, and Joe Apai. 2011. ``Terrorists brought down the plane!--No, actually it was a technical fault: processing corrections of emotive information.'' \emph{Quarterly journal of experimental psychology (2006)} 64(2): 283--310.

\leavevmode\hypertarget{ref-fernbach2019extreme}{}%
Fernbach, Philip M. et al. 2019. ``Extreme opponents of genetically modified foods know the least but think they know the most.'' \emph{Nature Human Behaviour} 3(3): 251--256.

\leavevmode\hypertarget{ref-flynn2017nature}{}%
Flynn, D. J., Brendan Nyhan, and Jason Reifler. 2017. ``The nature and origins of misperceptions: Understanding false and unsupported beliefs about politics.'' \emph{Political Psychology} 38: 127--150.

\leavevmode\hypertarget{ref-gaskell2010europeans}{}%
Gaskell, George et al. 2010. ``Europeans and Biotechnology in 2010. Winds of change?''

\leavevmode\hypertarget{ref-jerit2020political}{}%
Jerit, Jennifer, and Yangzi Zhao. 2020. ``Political misinformation.'' \emph{Annual Review of Political Science} 23: 77--94.

\leavevmode\hypertarget{ref-lazer2018science}{}%
Lazer, David M. J. et al. 2018. ``The science of fake news.'' \emph{Science} 359(6380): 1094--1096.

\leavevmode\hypertarget{ref-lewandowsky2012misinformation}{}%
Lewandowsky, Stephan et al. 2012. ``Misinformation and its correction: Continued influence and successful debiasing.'' \emph{Psychological Science in the Public Interest} 13(3): 106--131.

\leavevmode\hypertarget{ref-marcus2000affective}{}%
Marcus, George E., W. Russell Neuman, and Michael MacKuen. 2000. \emph{Affective intelligence and political judgment}. University of Chicago Press.

\leavevmode\hypertarget{ref-marcus2019applying}{}%
Marcus, George E. et al. 2019. ``Applying the theory of affective intelligence to support for authoritarian policies and parties.'' \emph{Political Psychology} 40: 109--139.

\leavevmode\hypertarget{ref-martel2020reliance}{}%
Martel, Cameron, Gordon Pennycook, and David G. Rand. 2020. ``Reliance on emotion promotes belief in fake news.'' \emph{Cognitive Research: Principles and Implications} 5(1): 47.

\leavevmode\hypertarget{ref-moors2013appraisal}{}%
Moors, Agnes et al. 2013. ``Appraisal theories of emotion: State of the art and future development.'' \emph{Emotion Review} 5(2): 119--124.

\leavevmode\hypertarget{ref-nyhan2019taking}{}%
Nyhan, Brendan et al. 2019. ``Taking fact-checks literally but not seriously? The effects of journalistic fact-checking on factual beliefs and candidate favorability.'' \emph{Political Behavior}: 1--22.

\leavevmode\hypertarget{ref-redlawsk2006feeling}{}%
Redlawsk, David. 2006. \emph{Feeling politics: Emotion in political information processing}. Springer.

\leavevmode\hypertarget{ref-roberts2018nobel}{}%
Roberts, Richard J. 2018. ``The Nobel Laureates' Campaign Supporting GMOs.'' \emph{Journal of Innovation \& Knowledge} 3(2): 61--65. \url{https://www.sciencedirect.com/science/article/pii/S2444569X18300064}.

\leavevmode\hypertarget{ref-rosenzweig2020misinformation}{}%
Rosenzweig, Leah et al. 2020. ``Misinformation and Emotions in Nigeria: The Case of COVID-19 Fake News.''

\leavevmode\hypertarget{ref-ryan2019actions}{}%
Ryan, Timothy J. 2019. ``Actions versus consequences in political arguments: Insights from moral psychology.'' \emph{The Journal of Politics} 81(2): 426--440.

\leavevmode\hypertarget{ref-ryan2017no}{}%
Ryan, Timothy J. 2017. ``No Compromise: Political Consequences of Moralized Attitudes.'' \emph{American Journal of Political Science} 61(2): 409--423.

\leavevmode\hypertarget{ref-ryan2014reconsidering}{}%
Ryan, Timothy J. 2014. ``Reconsidering moral issues in politics.'' \emph{The Journal of Politics} 76(2): 380--397.

\leavevmode\hypertarget{ref-sherman2002affective}{}%
Sherman, David K., and Heejung S. Kim. 2002. ``Affective perseverance: The resistance of affect to cognitive invalidation.'' \emph{Personality and Social Psychology Bulletin} 28(2): 224--237.

\leavevmode\hypertarget{ref-siegrist2020consumer}{}%
Siegrist, Michael, and Christina Hartmann. 2020. ``Consumer acceptance of novel food technologies.'' \emph{Nature Food} 1(6): 343--350.

\leavevmode\hypertarget{ref-skitka2020psychology}{}%
Skitka, Linda J. et al. 2020. ``The Psychology of Moral Conviction.'' \emph{Annual Review of Psychology} 72.

\leavevmode\hypertarget{ref-skitka2018attitude}{}%
Skitka, Linda J., Daniel C. Wisneski, and Mark J. Brandt. 2018. ``Attitude moralization: Probably not intuitive or rooted in perceptions of harm.'' \emph{Current Directions in Psychological Science} 27(1): 9--13.

\leavevmode\hypertarget{ref-swire2019they}{}%
Swire-Thompson, Briony et al. 2019. ``They Might Be a Liar But They're My Liar: Source Evaluation and the Prevalence of Misinformation.'' \emph{Political Psychology}.

\leavevmode\hypertarget{ref-thorson2016belief}{}%
Thorson, Emily. 2016. ``Belief Echoes: The Persistent Effects of Corrected Misinformation.'' \emph{Political Communication} 33(3): 460--480.

\leavevmode\hypertarget{ref-vosoughi2018spread}{}%
Vosoughi, Soroush, Deb Roy, and Sinan Aral. 2018. ``The spread of true and false news online.'' \emph{Science} 359(6380): 1146--1151.

\leavevmode\hypertarget{ref-walter2020fact}{}%
Walter, Nathan et al. 2020. ``Fact-Checking: A Meta-Analysis of What Works and for Whom.'' \emph{Political Communication} 37(3): 350--375.

\leavevmode\hypertarget{ref-wardle2017information}{}%
Wardle, Claire, and Hossein Derakhshan. 2017. ``Information Disorder: Toward an interdisciplinary framework for research and policy making.'' \emph{Council of Europe Report} 27.

\leavevmode\hypertarget{ref-weeks2015emotions}{}%
Weeks, Brian E. 2015. ``Emotions, Partisanship, and Misperceptions: How Anger and Anxiety Moderate the Effect of Partisan Bias on Susceptibility to Political Misinformation.'' \emph{Journal of Communication} 65(4): 699--719.

\leavevmode\hypertarget{ref-widmann2021how}{}%
Widmann, Tobias. 2021. ``How Emotional Are Populists Really? Factors Explaining Emotional Appeals in the Communication of Political Parties.'' \emph{Political Psychology} 42(1): 163--181.

\leavevmode\hypertarget{ref-wisneski2017moralization}{}%
Wisneski, Daniel C., and Linda J. Skitka. 2017. ``Moralization through moral shock: Exploring emotional antecedents to moral conviction.'' \emph{Personality and Social Psychology Bulletin} 43(2): 139--150.

\end{document}

### 2021-09-09 data-preparation.R

@@ -1,0 +1,177 @@
+# load packages
+library(tidyverse)
+library(sjmisc)
+library(scales)
+
+rm(list = ls())
+
+# load data (so far pre-test data)
+#load("Z:/3_current_Pretest/Pretest_W53.RData")
+
+# internal release
+load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+
+data <- GIP_W53_V1
+names(data)
+
+
+# pre-attitude
+frq(data$CE53482)
+
+data$pre_attitude <- ifelse(data$CE53482 < 0, NA, data$CE53482)
+frq(data$pre_attitude)
+
+# random assignment
+frq(data$expCE53483)
+
+data$exp_group <- fct_drop(recode(data$expCE53483,
+                         "1. Gruppe 1" = "control",
+                         "2. Gruppe 2" = "neutral-correction",
+                         "3. Gruppe 3" = "counter-frame",
+                         "4. Gruppe 4" = "low-misinfo",
+                         "5. Gruppe 5" = "low-misinfo-neutral-correction",
+                         "6. Gruppe 6" = "low-misinfo-counter-frame",
+                         "7. Gruppe 7" = "high-misinfo",
+                         "8. Gruppe 8" = "high-misinfo-neutral-correction",
+                         "9. Gruppe 9" = "high-misinfo-counter-frame"))
+frq(data$exp_group)
+
+# exclude cases with no random assignment
+data <- data[complete.cases(data$exp_group),]
+
+data <- data %>% 
+  mutate(correction = case_when(
+    exp_group == "control" | exp_group == "low-misinfo" | exp_group == "high-misinfo" ~ "none",
+    exp_group == "neutral-correction" | exp_group == "low-misinfo-neutral-correction" | exp_group == "high-misinfo-neutral-correction" ~ "neutral",
+    exp_group == "counter-frame" | exp_group == "low-misinfo-counter-frame" | exp_group == "high-misinfo-counter-frame" ~ "counterframe"
+  ) %>% 
+    fct_relevel("none", "neutral", "counterframe"),
+  misinfo = case_when(
+    exp_group == "control" | exp_group == "neutral-correction" | exp_group == "counter-frame" ~ "none",
[…]

### 2021-09-10 paper.Rmd
include link to citation style file

@@
 bibliography: references.bib
-csl: american-journal-of-political-science.csl
+csl: american-journal-of-political-science.csl # https://github.com/citation-style-language/styles
 output:

### 2021-09-10 paper.Rmd
references from citavi or google drive

--- 2021-09-09 paper.Rmd
@@
 
-bibliography: references.bib
-csl: american-journal-of-political-science.csl
+bibliography: references.bib # to be downloaded from citavi or google drive
+csl: american-journal-of-political-science.csl # https://github.com/citation-style-language/styles
 output:

### 2021-09-10 data-preparation.R
data later publicly available

@@ -10,6 +10,7 @@ rm(list = ls())
 
 # internal release
 load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+# later, the data will be made public
 
 data <- GIP_W53_V1
 names(data)

### 2021-09-13 paper.Rmd
update

@@
 date: \today
-linestretch: 1.2
+linestretch: 1.5
 colorlinks: true
@@
 library(scales)
+library(knitr)
 ```
@@
 <!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
-While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2020misinformation; @weeks2015emotions].
+While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions].
 This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information.
@@
 
-## Misinformation and corrective information
+## Misinformation and corrective information {-}
 
@@
 
-## Emotions and political learning
+## Emotions, political learning, moralization {-}
 
@@
 
-## Hypotheses
+## Empirical expectations {-}
 
@@
 
-## GMO - a case for emotional reactions and misinformation
+## GMO - a case for emotional reactions and misinformation {-}
 
@@
 
+To investigate the direct effects of misinfomation and corrective information conditional on
+
+The study will be conducted as a survey experiment embedded within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
+The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
+%The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample.
+Households that did not posses the technical devices to participate online have been equipped with such devices.
+Survey waves are conducted on a bi-monthly basis, the fieldwork for this study will take place in May 2021.
+Because data collection is done by a third party, I do not have control over the sample size. From results of previous waves I expect a sample size of approximately 4,400 respondents.
+Fieldwork will be conducted in May 2021.
+Preliminary data for my own analyses will be available in July 2021 and a scientific use file will be made publicly available at GESIS data archive in December 2021.
+Pre-registered at OSF: https://osf.io/y63xu
+
+## Treatment Material
+
+## Outcome measures
+
 # Results
-
-## Descriptives
 
@@
 
-```{r, warning=FALSE, echo=FALSE}
+## Descriptive statistics {-}
+
+Table \@ref(tab:descriptives) shows
+
+```{r descriptives, warning=FALSE, echo=FALSE, fig.pos="H"}
 # only control group
@@
 
-datasummary_skim(descriptives_control)
+knitr::kable(datasummary_skim(descriptives_control, histogram = FALSE, output = "data.frame"), 
+             caption = "Descriptive statistics")
 
@@
 
-## Moralization
+## Moralization through misinformation {-}
+
+To quantify the effects of the different treatment conditions on the outcomes of interest, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.
 
@@
 
-```{r, echo = FALSE}
+```{r fig-1, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 # H1: misinformation effect
@@
 
-```{r, echo = FALSE}
+## Effectiveness of neutral correction {-}
+
+```{r fig-2, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 #sjPlot::save_plot("./figures/p_neutral_correction.png", width = 24, height = 16)
+```
+
+## Counter-frame {-}
+
+```{r fig-3, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 
+```
+
+# Conclusion
+
+Summary:
+
+* Persuasive, emotionally framed misinformation can increase moralization of opinions about GM food.
+* Effects persist with neutral corrective information, but counter-framing works.
+* Overall small effects: Explained variance ($R^2$) of $7 \%$ for beliefs, $3.5 \%$ for opinions, $1 \%$ for moral conviction.
+
+Limitations:
+
+* is it really emotions or "just" the pictures?
+
+Implications:
+
+* When misinformation is emotionally charged, fact-checkers may have to think about counter-frames to eliminate the misinformation effect.
+* a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.
+genetic engineering provides avenues with large potential benefits @roberts2018nobel, @planck2019scientists, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
+* Similar effects possible on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.
+
+\newpage
+
+# References
+
+\linespread{1}
+\setlength{\parindent}{-0.2in}
+\setlength{\leftskip}{0.2in}
+\setlength{\parskip}{8pt}
+\noindent
+
+::: {#refs}
+
+:::
+
+\newpage
+
+# Appendix {-}
+
+```{r fig-4, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+
 # Appendix
@@
                 size = 0.5,
-                width = 0.5,
+                width = 0.25,
                 position = position_dodge(width = 0.5)) +
@@
        y = "") +
-  facet_wrap(~fct_rev(type), scales = "free_y") +
+  facet_wrap(group~fct_rev(type), scales = "free_y") +
   coord_flip() +
@@
         legend.position = "bottom",
-        legend.key.size = unit(1, 'cm')) +
+        legend.key.size = unit(1, 'cm'),
+        strip.text.x = element_text(size = 8)
+  ) +
   guides(color = guide_legend(reverse = TRUE),
@@
 ```
-
-# Conclusion
-
-\newpage
-
-# References
-
-\linespread{1}
-\setlength{\parindent}{-0.2in}
-\setlength{\leftskip}{0.2in}
-\setlength{\parskip}{8pt}
-\noindent
-
-::: {#refs}
-
-:::

### 2021-09-14 data-preparation.R
include check whether data already in workspace

@@ -9,7 +9,10 @@ rm(list = ls())
 #load("Z:/3_current_Pretest/Pretest_W53.RData")
 
 # internal release
-load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+if(!exists("GIP_W53_V1")) {
+  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+}
+
 # later, the data will be made public
 
 data <- GIP_W53_V1

### 2021-09-14 paper.Rmd
various mostly formal additions

@@
 ---
-title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science for excellent comments and helpful suggestions on this study.
+title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
 This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
@@
 \vspace{.8cm}
+# abstract-template: Everyone agrees that this issue is really important.
+But we do not know much about this specific question, although it matters a great deal, for these reasons.
+We approach the problem from this perspective.
+Our research design focuses on these cases and relies on these data, which we analyze using this method.
+Results show what we have learned about the question.
+They have these broader implications.
 
@@
 <!-- % emotion in political science -->
-Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
 In this study, I follow recent work in political science inspired by moral psychology.
@@
 Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude].
-While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
 
@@
 Specifically, it provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims.
-
-<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
-The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, hasell2020differential, carnahan2019processing].
-This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans].
-GM food is also a topic on which there are misperceptions, possibly due to disinformation.
-Of course, there is legitimate criticism on the use of genetic modification.
-However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].
 
@@
 
+Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
+
 Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
@@
 A study has shown that moral shock works to moralize attitudes towards GM food.
+
+While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
+
 However, the main part of the material used there can be classified as misinformation
@@
 
+<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
+The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, hasell2020differential, carnahan2019processing].
+This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans].
+GM food is also a topic on which there are misperceptions, possibly due to disinformation.
+Of course, there is legitimate criticism on the use of genetic modification.
+However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].
+
 # Data & Methods
 
-To investigate the direct effects of misinfomation and corrective information conditional on
+To investigate the direct effects of misinformation and corrective information conditional on
 
@@
 
-## Treatment Material
+## Treatment Material {-}
 
-## Outcome measures
+The experimental treatment consists of a combination of two types of information.
+One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats.
+The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional.
+The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed.
+The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128].
+If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.
+
+The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon.
+This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information @[lewandowsky2020debunking].
+As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format.
+By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food.
+This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.
+
+Assignment is done by simple random assignment into equal sized groups.
+Table gives an overview over the nine resulting experimental groups.
+The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned).
+For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' @[kuklinski2000misinformation, p.805].
+
+## Outcome measures {-}
+
+As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct".
+
+ To provide a broad measure of policy support, I use a measure by @[diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".
+
+I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" @[skitka2020psychology].
+This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them.
+Both items will be combined into a simple additive index.
+
+I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food.
+Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds @[skitka2018attitude].
+The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".
+
+## Ethics {-}
+
+An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
+Because all shown information has in a similar format been in the media, there is no deception taking place.
+The experimental treatment intends to evoke emotions, but I consider them within reason.
+There are two main reasons for that.
+First, the material including the used imagery has been used in a previously published study @[clifford2019emotional].
+Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
+The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents.
+Because this approval has been granted, this study did not require further approval by an ethics committee.
 
 # Results
+
+This paper is written in RMarkdown based on a template available here: https://osf.io/q395s
 
@@
 
-Table \@ref(tab:descriptives) shows
+Table \@ref(tab:descriptives)
 
-```{r descriptives, warning=FALSE, echo=FALSE, fig.pos="H"}
+```{r descriptives, echo=FALSE, message=FALSE, warning=FALSE}
+library(modelsummary)
 # only control group
 
-descriptives_control <- data %>% 
+data %>% 
   filter(exp_group == "control") %>% 
@@
          "Policy restrictions" = "policy_opinion",
-         "Moral conviction" = "moral_conviction")
-
-knitr::kable(datasummary_skim(descriptives_control, histogram = FALSE, output = "data.frame"), 
-             caption = "Descriptive statistics")
+         "Moral conviction" = "moral_conviction") %>% 
+  datasummary_skim(histogram =FALSE,
+                   output = "default",
+                   title = "Descriptive statistics in the control group")
 
@@
 \newpage
+\pagenumbering{roman}
 
 # Appendix {-}
+
+## Information effect
 
@@
 ```
+
+\newpage
+
+## Regression tables
+
+```{r reg-output-emotion, message=FALSE, warning=FALSE, echo = FALSE}
+
+library(modelsummary)
+
+cm <- c('(Intercept)'    = 'Control',
+        'exp_grouplow-misinfo'    = 'Low emotion misinf.',
+        'exp_grouphigh-misinfo' = 'High emotion misinf.',
+        'exp_grouplow-misinfo-neutral-correction' = 'Neutral correction to low emotion misinf.',
+        'exp_grouphigh-misinfo-neutral-correction' = 'Neutral correction to high emotion misinf.',
+        'exp_grouplow-misinfo-counter-frame' = 'Counterframe low emotion misinf.',
+        'exp_grouphigh-misinfo-counter-frame' = 'Counterframe to high emotion misinf.',
+        'exp_groupneutral-correction'    = 'Neutral correction w/o misinf.',
+        'exp_groupcounter-frame'    = 'Counterframe w/o misinf.')
+
+models_emotions <- list("Anger" = lm_emotion_anger,
+                        "Anxiety" =  lm_emotion_anxiety,
+                        "Disgust" = lm_emotion_disgust,
+                        "Sadness" = lm_emotion_sadness)
+
+modelsummary(models_emotions,
+             coef_map = cm,
+             title = 'Regression table for emotions.', 
+             stars = c('*' = .05, '**' = .01, '***' = 0.01),
+             gof_omit = 'DF|Deviance|F|Log|AIC|BIC',
+             output = 'markdown')
+
+```
+\newpage
+
+```{r reg-output-belief-opinion, message=FALSE, warning=FALSE, echo = FALSE}
+
+models <- list("Belief GMO safety" = lm_knowledge_safe,
+               "Policy restrictions" =  lm_policy_opinion,
+               "Moral conviction" = lm_moral_conviction)
+
+modelsummary(models,
+             coef_map = cm,
+             title = 'Regression table for belief and policy opinion.', 
+             stars = c('*' = .05, '**' = .01, '***' = 0.01),
+             gof_omit = 'DF|Deviance|F|Log|AIC|BIC',
+             output = 'markdown')
+
+```
+
+\newpage
+
+## Treatment material
+
+```{r screenshot-low-misinfo, fig.align = 'center', out.width = "100%", fig.cap = "Low emotion misinformation", echo = FALSE}
+knitr::include_graphics("./figures/misinfo_low_emotion.png")
+```
+
+```{r screenshot-high-misinfo, fig.align = 'center', out.width = "100%", fig.cap = "High emotion misinformation", echo = FALSE}
+knitr::include_graphics("./figures/misinfo_high_emotion.png")
+```
+
+```{r screenshot-neutral-correction, fig.align = 'center', out.width = "100%", fig.cap = "Neutral correction", echo = FALSE}
+knitr::include_graphics("./figures/neutral_correction.png")
+```
+
+```{r screenshot-counterframe, fig.align = 'center', out.width = "100%", fig.cap = "Counterframe correction", echo = FALSE}
+knitr::include_graphics("./figures/counter_frame.png")
+```

### 2021-09-14 paper.Rmd
various mostly formal additions

@@
 
-The experimental treatment consists of a combination of two types of information.
-One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats.
-The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional.
-The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed.
-The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128].
-If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.
-
-The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon.
-This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information @[lewandowsky2020debunking].
-As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format.
-By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food.
-This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.
-
-Assignment is done by simple random assignment into equal sized groups.
-Table gives an overview over the nine resulting experimental groups.
-The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned).
-For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' @[kuklinski2000misinformation, p.805].
-
 ## Outcome measures {-}
-
-As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct".
-
- To provide a broad measure of policy support, I use a measure by @[diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".
-
-I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" @[skitka2020psychology].
-This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them.
-Both items will be combined into a simple additive index.
-
-I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food.
-Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds @[skitka2018attitude].
-The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".
-
-## Ethics {-}
-
-An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
-Because all shown information has in a similar format been in the media, there is no deception taking place.
-The experimental treatment intends to evoke emotions, but I consider them within reason.
-There are two main reasons for that.
-First, the material including the used imagery has been used in a previously published study @[clifford2019emotional].
-Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
-The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents.
-Because this approval has been granted, this study did not require further approval by an ethics committee.
 

### 2021-09-15 paper.Rmd
some progress

--- 2021-09-14 paper.Rmd
@@
 ---
-title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
-This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
+title: |
+  | Emotional framing and the effectiveness of 
+  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
+This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
+  | *Working Paper*
 author: |
+  | *Working paper*
+  |
   | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
@@
 ---
+
+<!-- This paper is written in RMarkdown based on a template available here: https://osf.io/q395s -->
 
@@
 
-To investigate the direct effects of misinformation and corrective information conditional on
+To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
+The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
+The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample.
+Households that did not posses the technical devices to participate online have been equipped with such devices.
+Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.
 
-The study will be conducted as a survey experiment embedded within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
-The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
-%The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample.
-Households that did not posses the technical devices to participate online have been equipped with such devices.
-Survey waves are conducted on a bi-monthly basis, the fieldwork for this study will take place in May 2021.
-Because data collection is done by a third party, I do not have control over the sample size. From results of previous waves I expect a sample size of approximately 4,400 respondents.
-Fieldwork will be conducted in May 2021.
-Preliminary data for my own analyses will be available in July 2021 and a scientific use file will be made publicly available at GESIS data archive in December 2021.
-Pre-registered at OSF: https://osf.io/y63xu
+The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction).
+Respondents were randomly assigned to one of the 9 resulting groups, thus yielding 8 treatment conditions and one control condition were no information treatment was shown.
+
+An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
+Because all shown information has in a similar format been in the media, there is no deception taking place.
+The experimental treatment intends to evoke emotions, but I consider them within reason.
+There are two main reasons for that.
+First, the material including the used imagery has been used in a previously published study @[clifford2019emotional].
+Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
+The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents.
+Because this approval has been granted, this study did not require further approval by an ethics committee.
 
@@
 
-## Ethics {-}
-
-An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
-Because all shown information has in a similar format been in the media, there is no deception taking place.
-The experimental treatment intends to evoke emotions, but I consider them within reason.
-There are two main reasons for that.
-First, the material including the used imagery has been used in a previously published study @[clifford2019emotional].
-Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
-The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents.
-Because this approval has been granted, this study did not require further approval by an ethics committee.
-
 # Results
-
-This paper is written in RMarkdown based on a template available here: https://osf.io/q395s
 

### 2021-09-15 paper.Rmd
some progress

@@
 This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
-  | *Working Paper*
 author: |
@@
 
-To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
+To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
 The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
@@
 The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction).
-Respondents were randomly assigned to one of the 9 resulting groups, thus yielding 8 treatment conditions and one control condition were no information treatment was shown.
+Respondents were randomly assigned to one of the nine resulting groups, thus yielding eight treatment conditions and one control condition were no information treatment was shown.
 
@@
 
-```{r, warning=FALSE, echo=FALSE}
+```{r, warning=FALSE, echo=FALSE, message = FALSE}
 # load data
@@
 
-Table \@ref(tab:descriptives)
+Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects.
+On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint.
+Anxiety was above the midpoint with a mean of 2.4 whereas disgust was realtively low with a mean of 1.3, and sadness again moderatly below the midpoint with 1.7.
+These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.
+
+With regard to the factual belief that GM food is safe to eat, the sample is skewed toward the view that this is rather not true (mean = 3.2 on a scale from 1 to 7).
+When it comes to opinions about policy restrictions for GM food, respondents are rather in favor (mean = 5.2 on a scale from 1 to 7).
+Finally, the measure of moral conviction is close to the midpoint (mean = 4.4 on a scale from 0 to 8).
 
@@
 
-To quantify the effects of the different treatment conditions on the outcomes of interest, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.
+I now turn to the effects of the informational treatments on the outcome variables.
+To quantify these effects, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.
+For a more intuitive interpretation of the effect sizes - and to be able to compare outcome variables measured on different scales - I rescale all outcome variables to range from 0 to 1.
+I do not include any control variables because the random assignment of the experimental design provides the better safeguard against any possible confounders.
+
+In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.
 
@@
 
-```{r fig-1, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message.
+The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message.
+The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
+
+```{r misinfo-only, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 # H1: misinformation effect
@@
   mutate(emotion = ifelse(str_detect(group, "low-misinfo"), 
-                          "Low emotional misinformation", "High emotional misinformation"))
+                          "Low emotion misinformation", "High emotion misinformation"))
 
@@
         legend.position = "bottom",
-        legend.key.size = unit(1, 'cm'),
+        legend.key.size = unit(0.1, 'cm'),
         strip.text.x = element_text(size = 8)
@@
 pdata_neutral_correction <- pdata_neutral_correction %>% 
-  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
+  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotion misinformation", "High emotion misinformation"),
          correction = ifelse(str_detect(group, "correction"), "Neutral correction", "No correction"))
@@
 pdata_counter_frame <- pdata_counter_frame %>% 
-  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
-         correction = ifelse(str_detect(group, "counter"), "Counter-frame correction", "No correction"))
+  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotion misinformation", "High emotion misinformation"),
+         correction = ifelse(str_detect(group, "counter"), "Counterframe correction", "No correction"))
 

### 2021-09-15 data-preparation.R
include possibility to load data from folder

@@ -2,11 +2,14 @@
 library(tidyverse)
 library(sjmisc)
 library(scales)
+library(here)
 
 rm(list = ls())
 
 # load data (so far pre-test data)
 #load("Z:/3_current_Pretest/Pretest_W53.RData")
+
+load(here("GIP_W53_V1.RData"))
 
 # internal release
 if(!exists("GIP_W53_V1")) {

### 2021-09-15 paper.Rmd
bullet points for theory

@@
 Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude]. 
+While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
 
@@
 <!-- % Final Punchline on broader relevance -->
-In the view of many scientists [@roberts2018nobel, planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
+In the view of many scientists [@roberts2018nobel, @planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
 In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.
 
+\newpage
 # Theory
@@
 ## Misinformation and corrective information {-}
+
+*t.b.d.*
+
+* Definition and distinction between misinformation and disinformation.
+* Some worries about backfire effects, but usually, corrective information works, and guidelines for best practice corrections have been collected.
+* Emotions may contribute to the spread of false information.
+* Relatively little evidence on the role of emotions for misinformation and corrective information.
 
@@
 
-Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information.
+<!-- Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information. -->
 
-The effects on related opinions are much less clear.
-Some studies find effects, some not.
+<!-- The effects on related opinions are much less clear. -->
+<!-- Some studies find effects, some not. -->
 
-Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected.
+<!-- Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected. -->
 
-## Emotions, political learning, moralization {-}
+## Emotional framing and moralization {-}
+
+*t.b.d.*
+
+* Emotions feature prominently in political communication.
+* Framing.
+* Some known effects on information processing.
+* Findings from moral psychology on moralization through disgust.
+* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
 
@@
 
-Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
+<!-- Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
+-->
 
-Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
+<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
+-->
 
-Research has shown that... anger... anxiety... enthusiasms...
+<!-- Research has shown that... anger... anxiety... enthusiasms... -->
 
-A recently growing interest is in disgust.
-Connection to morality, relevant for immigration, homeless ...
+<!-- A recently growing interest is in disgust. -->
+<!-- Connection to morality, relevant for immigration, homeless ... -->
 
-A study has shown that moral shock works to moralize attitudes towards GM food.
+<!-- A study has shown that moral shock works to moralize attitudes towards GM food. -->
 
-While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
-
-However, the main part of the material used there can be classified as misinformation
-GM food is an issue on which disinformation exists.
+<!-- However, the main part of the material used there can be classified as misinformation -->
+<!-- GM food is an issue on which disinformation exists.
 This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence.
 Least knowledgeable people are the fiercest opponents.
-So
+So -->
 
@@
 
-## GMO - a case for emotional reactions and misinformation {-}
+## GM food - a case for emotional reactions and misinformation {-}
 
 <!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
-The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, hasell2020differential, carnahan2019processing].
+The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, @hasell2020differential, @carnahan2019processing].
 This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans].
@@
 There are two main reasons for that.
-First, the material including the used imagery has been used in a previously published study @[clifford2019emotional].
+First, the material including the used imagery has been used in a previously published study [@clifford2019emotional].
 Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
@@
 The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon.
-This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information @[lewandowsky2020debunking].
+This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information [@lewandowsky2020debunking].
 As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format.
@@
 The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned).
-For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' @[kuklinski2000misinformation, p.805].
+For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' [@kuklinski2000misinformation, p.805].
 
@@
 
-To provide a broad measure of policy support, I use a measure by @[diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".
+To provide a broad measure of policy support, I use a measure by [@diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".
 
-I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" @[skitka2020psychology].
+I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" [@skitka2020psychology].
 This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them.
@@
 I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food.
-Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds @[skitka2018attitude].
+Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds [@skitka2018attitude].
 The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".
@@
 On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint.
-Anxiety was above the midpoint with a mean of 2.4 whereas disgust was realtively low with a mean of 1.3, and sadness again moderatly below the midpoint with 1.7.
+Anxiety was above the midpoint with a mean of 2.4 whereas disgust was relatively low with a mean of 1.3, and sadness again moderately below the midpoint with 1.7.
 These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.
@@
 In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.
+Without any controls, the coefficients refer to the difference in the mean of the outcome between the control and the respective treatment group.
+^[The significance of the coefficients is equivalent to running separate t-tests between the group means, which raises the issue of adjusting for multiple testing.
+There are different approaches from different traditions to that.
+When conducting many independent t-tests, it seems straight-forward to adjust for multiple testing, although the general advise to always adjust for multiple testing is not warranted [@althouse2016adjust].
+Another approach is to conduct ANOVA, which is an omnibus test, with the null-hypothesis of no effect of the treatment conditions together versus the control.
+To check for individual differences between experimental groups, post-hoc tests can be conducted with respective appreciation of multiple testing adjustments.
+This is appropriate if one has no specific hypotheses to begin with.
+If one has these hypotheses, however, one can use ANOVA with planned contrasts (e.g.
+[@field2012discovering]), which is then again equivalent to the results of a regression with dummy variables coded according to the specified contrasts.
+When I use multiple outcomes in my pre-registered hypotheses, the expectation is not that any of the outcomes shows an effect (in which case adjustment for multiple testing should be done), but rather a joint pattern of effects.]
 
@@
 
-## Counter-frame {-}
+## Counterframe correction {-}
 
@@
 
+\newpage
 # Conclusion
+
+*t.b.d.*
 
@@
 
-* is it really emotions or "just" the pictures?
+* No distinction between emotional framing and visual cues.
 
@@
 
-* When misinformation is emotionally charged, fact-checkers may have to think about counter-frames to eliminate the misinformation effect.
+* When misinformation is emotionally charged, fact-checkers may have to think about counterframes to eliminate the misinformation effect.
 * a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.

### 2021-09-16 paper.Rmd
writing

@@
   | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
-This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
+This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on "The Political Economy of Reforms" (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
 author: |
@@
 <!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
-While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions].
+While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored.
 This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information.
 We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread].
-In general, moral-emotional language is rewarded on social media [@brady2017emotion] and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how].
+In general, moral-emotional language is rewarded on social media [@brady2017emotion].
+<!-- and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how]. -->
 
@@
 However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer.
-Such a piece of information can be classified as misinformation in the sense of going against ``best available evidence in the public sphere''[@flynn2017nature, p.128].
+Such a piece of information can be classified as misinformation in the sense of going against "best available evidence in the public sphere"[@flynn2017nature, p.128].
 From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.
@@
 
-* Definition and distinction between misinformation and disinformation.
-* Some worries about backfire effects, but usually, corrective information works, and guidelines for best practice corrections have been collected.
-* Emotions may contribute to the spread of false information.
-* Relatively little evidence on the role of emotions for misinformation and corrective information.
-
-<!-- % some definitions from Kuklinski and Flynn to Jerit -->
-
-<!-- % information corrects factual beliefs, but the effects of information on attitudes are mixed -->
-
-<!-- Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information. -->
-
-<!-- The effects on related opinions are much less clear. -->
-<!-- Some studies find effects, some not. -->
-
-<!-- Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected. -->
+* Definition of related terms: misinformation, disinformation, fake news, misperceptions [@flynn2017nature; @jerit2020political; @kuklinski2000misinformation].
+* Wide-spread misperceptions may distort public opinion and corrective information (fact-checking) is one way to tackle the problem.
+* People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
+* However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
+* An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
+* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]).
 
@@
 
-* Emotions feature prominently in political communication.
-* Framing.
-* Some known effects on information processing.
+* Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
 * Findings from moral psychology on moralization through disgust.
 * Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
+* Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.
+
+<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
+-->
 
@@
 -->
-
-<!-- Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
--->
-
-<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
--->
-
-<!-- Research has shown that... anger... anxiety... enthusiasms... -->
-
-<!-- A recently growing interest is in disgust. -->
-<!-- Connection to morality, relevant for immigration, homeless ... -->
-
-<!-- A study has shown that moral shock works to moralize attitudes towards GM food. -->
-
-<!-- However, the main part of the material used there can be classified as misinformation -->
-<!-- GM food is an issue on which disinformation exists.
-This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence.
-Least knowledgeable people are the fiercest opponents.
-So -->
 
@@
 
-Together, hypotheses 2a and 2b represent the existence of a ``belief echo'', i.e.
+Together, hypotheses 2a and 2b represent the existence of a "belief echo", i.e.
 the continued effect of a piece of misinformation on attitudinal or emotional outcomes, despite successful factual correction.
@@
 If this is the case, a more effective strategy may be to counter the emotional misinformation not with a neutral correction but with a correction embedded in an emotional counter-frame.
-If equally persuasive, such counterframes have been shown to negate the effect of an original frame [@chong2013counterframing].
+If equally persuasive, such counterframes have been shown to negate the effect of an original frame [@chong2007competitive; @chong2013counterframing].
 I expect a similar effect to hold in response to emotionally framed misinformation.
 
-* H3 A correction embedded in a persuasive emotional counter-frame eliminates the misinformation effect on policy opinions, moral conviction, and emotions.
+* H3 A correction embedded in a persuasive emotional counterframe eliminates the misinformation effect on policy opinions, moral conviction, and emotions.
 
@@
 
-<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
-The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, @hasell2020differential, @carnahan2019processing].
+<!-- % explain the relevance of GMO as an issue, also in relation to moralization -->
+The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does; @hasell2020differential; @carnahan2019processing].
 This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans].
@@
 
-To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
+To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[The pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
 The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
-The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample.
-Households that did not posses the technical devices to participate online have been equipped with such devices.
 Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.
 
-The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction).
+The between-subjects survey experiment followed a 3 $\times$ 3 design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction).
 Respondents were randomly assigned to one of the nine resulting groups, thus yielding eight treatment conditions and one control condition were no information treatment was shown.
 
-An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
-Because all shown information has in a similar format been in the media, there is no deception taking place.
-The experimental treatment intends to evoke emotions, but I consider them within reason.
-There are two main reasons for that.
-First, the material including the used imagery has been used in a previously published study [@clifford2019emotional].
-Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
-The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents.
-Because this approval has been granted, this study did not require further approval by an ethics committee.
+At the end of the survey, respondents received an extensive debriefing clarifying the sources of all information that was shown and on the intention of the experiment.
+Because all shown information has in a similar format been in the media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt], there is no deception taking place.
+The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is a pleasant experience for respondents.
 
@@
 
-The experimental treatment consists of a combination of two types of information.
-One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats.
+The experimental treatment consists of a combination of two types of information, each mimicking a short online news article.
+One type of information reports the results of a study claiming that genetically modified corn lead to increased cancer rates in rats, and features the headline *"GM corn-fed rats die earlier"*.
 The wording is partly taken from actual news coverage of the study in 2012, and partly from treatment material in @clifford2019emotional.
-The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed.
+The respective study has been retracted due to non-adherence to scientific procedures, and the current scientific evidence does not point to severe health problems associated with genetically modified animal feed [@brody2018are].
 The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128].
@@
 As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format.
+The headline for the neutral correction reads *"Genetic engineering does not introduce novel risks"*.
 By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food.
-This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.
+This is done in a further condition, where the corrective information is embedded in a counterframe, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.
+Here, the headline reads *"Genetic research for a more sustainable agriculture"*.
+^[Screenshots of the shown information treatments in the German original are included in the Appendix in figures \@ref(fig:screenshot-low-misinfo) to \@ref(fig:screenshot-counterframe).]
 
-Assignment is done by simple random assignment into equal sized groups.
-Table gives an overview over the nine resulting experimental groups.
 The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned).
-For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' [@kuklinski2000misinformation, p.805].
+For respondents, the manipulation is thus designed in a way that both types of information "hits them between the eyes" [@kuklinski2000misinformation, p.805].
 
@@
 
+To acknowledge findings from the maturing corrective information literature on varying correction success for different outcomes, I include a set of several measures.
+
 As the factual belief of interest, I measure factual beliefs about the safety of GM foods by asking respondents whether they think that the statement that eating GM foods is safe for one's health is false or correct on a 7-point scale from "certainly false" to "certainly correct".
-
 To provide a broad measure of policy support, I use a measure by [@diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".
-
 I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" [@skitka2020psychology].
@@
 
-I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food.
+Lastly, I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food.
 Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds [@skitka2018attitude].
@@
 
-First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message.
+First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message.
 The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message.
-The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
+The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
 
-```{r misinfo-only, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome.
+Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group.
+For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.
+
+Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.
+
+```{r misinfo-only, fig.align="center", fig.cap="Effects of misinformation with low and high emotional framing without following correction.
+Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 # H1: misinformation effect
@@
 
-```{r fig-2, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
+
+* the neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation
+
+```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 
-```{r fig-3, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe).
+It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).
+
+```{r counterframe, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing.
+Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 
-* Persuasive, emotionally framed misinformation can increase moralization of opinions about GM food.
-* Effects persist with neutral corrective information, but counter-framing works.
+* Persuasive, emotionally framed misinformation can increase moralization of opinions about GM food, replicating findings from @clifford2019emotional.
+* A neutral correction is successful for low emotional misinformation, but effects of highly emotional misinformation persist with neutral corrective information
+* Providing a correction embedded in a positive counterframe message works for both low and high emotional misinformation.
 * Overall small effects: Explained variance ($R^2$) of $7 \%$ for beliefs, $3.5 \%$ for opinions, $1 \%$ for moral conviction.
@@
 
-* No distinction between emotional framing and visual cues.
+* No distinction between emotional framing and visual cues (see e.g. @hameleers2020picture).
 
@@
 
-* When misinformation is emotionally charged, fact-checkers may have to think about counterframes to eliminate the misinformation effect.
-* a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.
-genetic engineering provides avenues with large potential benefits @roberts2018nobel, @planck2019scientists, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
-* Similar effects possible on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.
+* When misinformation is emotionally charged, fact-checkers and science communicators may have to think about counterframes to eliminate the misinformation effect.
+* Concerning GM food, it is argued that a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].
+With emotional misinformation in circulation, persuasive counterframes seem necessary to even equal out their effects.
+* Similar effects seem likely on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.
 
@@
 
-## Information effect
+## Information effects
 
-```{r fig-4, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+```{r info-only, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation.
+Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 

### 2021-09-16 paper.Rmd
writing

@@
 <!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
-While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored.
+While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions].
 This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information.
@@
 * An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
-* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]).
+* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information.
 
@@
 * Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
+
 * Findings from moral psychology on moralization through disgust.
 * Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
-* Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.
 
-<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
--->
+Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
+
+A study has shown that moral shock works to moralize attitudes towards GM food.
+
+However, the main part of the material used there can be classified as misinformation
+GM food is an issue on which disinformation exists.
+This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence.
+Least knowledgeable people are the fiercest opponents.
+So
 
@@
 
-To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[The pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
+To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting].
 The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany.
@@
 
-First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message.
+First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message.
 The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message.
-The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
-
-In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome.
-Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group.
-For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.
-
-Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.
+The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
 
@@
 
-After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
-
-* the neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation
-
-```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+```{r fig-2, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 
-Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe).
-It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).
-
-```{r counterframe, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing.
+```{r fig-3, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing.
 Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
@@
 
-```{r info-only, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation.
+```{r fig-4, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation.
 Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

### 2021-09-16 paper.Rmd
set the date to 16. September (APSA upload ;-) )

--- 2021-09-16 paper.Rmd
@@
   | GESS and SFB 884, University of Mannheim
-date: \today
+date: September 16, 2021
 linestretch: 1.5

### 2021-09-16 paper.Rmd
back on track

@@
 <!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
+
 While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored.
@@
 In general, moral-emotional language is rewarded on social media [@brady2017emotion].
-<!-- and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how]. -->
 
@@
 <!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.
+-->
+
+<!-- % From Vasiloupolou / Wagner -->
+<!-- %Early research into emotional reactions either used one simple positive–negative dimension or a two-dimensional valence model (Marcus, 2003; Marcus et al., 2000).
+More recently, researchers have argued that distinct emotions may characterise individuals’ reactions (e.g.
+Huddy et al., 2007; Larsen and McGraw, 2011; Petersen, 2010; Smith et al., 2008; Valentino et al., 2008, 2011; Weber, 2012; see also Conover and Feldman, 1986).
+Appraisal theories of emotions suggest that various distinct emotions arise due to individuals’ cognitive interpretation of the situation they experience (e.g.
+Roseman, 1991; Smith and Ellsworth, 1985; for a review, see Brader and Marcus, 2013).
+Neuroscientific theories based on the neural processes that generate emotional responses have also increasingly turned towards a three-dimensional model of emotions that distinguishes between fear, anger and enthusiasm (Brader and Marcus, 2013; Lerner and Keltner, 2001).
+Research in political psychology building on both appraisal and neural process theories has therefore focused in particular on the causes and consequences of anger, fear1 and enthusiasm as distinct emotions (Carver and Harmon-Jones, 2009; Conover and Feldman, 1986; Huddy et al., 2005, 2007; Lerner and Keltner, 2000, 2001; Petersen, 2010; Smith and Ellsworth, 1985; Valentino et al., 2011).
+-->
+<!-- % -->
+<!-- %Applied to political phenomena, distinct emotions cause different types of actions and attitudes: fear leads to caution and openness, and anger to more confident and aggressive responses (Druckman and McDermott, 2008; MacKuen et al., 2010).
+For instance, whereas fear and anger both tend to increase intolerance and threat perceptions (e.g.
+Erişen and Kentmen-Cin, 2017), they lead to distinct ways of reacting to terrorist threats (Lerner et al., 2003).
+Fearful citizens are more wary of the increased risk and turn to isolationism, while angry citizens are less risk-averse and tend to favour pro-active intervention (Huddy et al., 2007).
+Anxious citizens are also more likely to seek out new information and process this carefully (Brader et al., 2008; Marcus et al., 2000; Merolla and Zechmeister, 2009; Valentino et al., 2008).
+Finally, enthusiasm increases the willingness to participate in politics, but like anger it also increases less careful forms of information-processing (Brader, 2006; Marcus et al., 2000).
 -->

### 2021-09-16 paper.Rmd
move ryan to theory
some more results

@@
 * Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
+While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
 * Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.
@@
 After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
-
-* the neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation
+The neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation.
+The emotional frame of the misinformation therefore moderates the correcion effectiveness.
+Although weakened, the effects on policy opinion and moral conviction induced by the high emotion misinformation persist after the correction, confirming H2b.
+Even if the correction significantly moves the factual belief after the high emotion misinformation, there still persists an effect of the misinformation relative to the control group.
+The correction is therefore not successful in eradicating the misinformation effect, which rejects H2a.
+In sum, there is no belief echo, but a moderation of the correction effectiveness.
 

### 2021-09-17 paper.Rmd
APSA contribution

@@
 author: |
-  | *Working paper*
+  | *work in progress*
   |
@@
 Substantively, I investigate the issue of genetically modified (GM) food.
-I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000).
-Results show what we have learned about the question.
-This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims.
+I implement a pre-registered survey experiment manipulating both misinformation and corrective information treatments within a panel study based on a probability sample of the general population in Germany.
+Results show, first, that an emotionally framed piece of misinformation can increase moralization, replicating previous findings.
+Second, a neutral correction to the misinformation is only effective if the misnformation does not include a strong emotional frame, whereas a correction embedded in a counterframe is effective for both low and high emotional misinformation.
 In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.
@@
 Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization].
-Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude].
-
+While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
+<!-- Moral conviction, i.e.
+the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude].
+-->
 <!-- % get more concrete linking belief echoes / affective perseverance and specific emotion with possible connection to moralization -->
-In his study, Clifford shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes.
+<!-- In his study, @clifford2019emotional shows that disgust-evoking messages can increase moral conviction with its consequences on polarizing attitudes.
 However, some of the treatment material used in his study comes from a later retracted study linking GMO and cancer.
 Such a piece of information can be classified as misinformation in the sense of going against "best available evidence in the public sphere"[@flynn2017nature, p.128].
-From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we might suspect that such emotional effects of frames may persist even in the face of successful factual correction.
+-->
 
 <!-- % Contribution bringing belief echoes and discrete emotions together -->
-This study contributes to the literature of corrective information by linking the concept of belief echoes to the perseverance of specific emotions with related consequences on attitudes.
-Specifically, it provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims.
+In the context of misinformation, it seems possible and worrisome that such an increase in moral conviction through emotional framing may also take place through a piece of factually incorrect information.
+If this is the case, how does the evoked information interfere with subsequent efforts to correct the misinformation?
+Do emotions evoked by a piece of misinformation make effective corrections less likely?
+And do such misinformation induced emotions and related attitudes persist despite successful correction?
+<!-- From findings of belief echoes [@thorson2016belief] and affective perseverance [@sherman2002affective] we can further suspect that such emotional effects of frames may persist even in the face of successful factual correction.
+-->
+<!-- This study contributes to the literature of corrective information by linking the concept of belief echoes to the perseverance of specific emotions with related consequences on attitudes.  -->
+Substantively, I provide evidence on the effect of emotionally charged misinformation on perceptions of GM food, and ways to effectively correct false claims.
 
+Results show, first, that an emotionally framed piece of misinformation can increase moralization, replicating previous findings.
+Second, a neutral correction to the misinformation is only effective if the misnformation does not include a strong emotional frame, whereas a correction embedded in a counterframe is effective for both low and high emotional misinformation.
 <!-- % Final Punchline on broader relevance -->
-In the view of many scientists [@roberts2018nobel, @planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
+<!-- In the view of many scientists [@roberts2018nobel, @planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
+-->
 In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.
 
-\newpage
 # Theory
@@
 * However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
-* An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
+* Especially troubling cases are "affective perseverance" [@sherman2002affective], and "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
 * Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]).
@@
 * Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
-* Findings from moral psychology on moralization through disgust.
-* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
-While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].
+* Findings from moral psychology on moralization through disgust [@clifford2019emotional; @wisneski2017moralization].
+* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition [@ryan2014reconsidering; @ryan2017no; @ryan2019actions].
 * Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.
@@
 
-From findings on moral conviction and how emotional frames induce it [@clifford2019emotional; @wisneski2017moralization], I expect that a persuasive message evoking negative emotions, notably disgust and anger, can increase perceptions of the respective issue as being one of moral conviction.
+From findings on moral conviction and how emotional frames induce it [@clifford2019emotional; @wisneski2017moralization], I expect that a persuasive piece of misinformation evoking negative emotions, notably disgust and anger, can increase perceptions of the respective issue as being one of moral conviction.
 In order to count as a persuasive message, I expect the message to affect both factual beliefs and policy opinions in its direction.
 
-* H1 A persuasive, emotionally framed message evoking negative emotions of disgust and anger makes individuals more likely to favor policy restrictions on the respective issue and to perceive the issue as one of moral conviction.
+* H1 A persuasive, emotionally framed misinformation evoking negative emotions of disgust and anger makes individuals more likely to favor policy restrictions on the respective issue and to perceive the issue as one of moral conviction.
 
@@
 
-* H2a Corrective information is effective in eliminating the misinformation effect on factual beliefs.
+* H2a Corrective information is effective in eliminating the misinformation effect on factual beliefs, irrespective of the emotional framing of the misinformation.
 
@@
 If only H2a is supported, the correction is altogether successful.
+If H2a is not supported, the correction is either completely ineffective or its effectiveness is moderated by the emotional framing of the misinformation.
 
@@
 Finally, the measure of moral conviction is close to the midpoint (mean = 4.4 on a scale from 0 to 8).
+These results indicate that the misperception that GM food is not safe for one own's health is wide-spread in Germany and that policy restrictions are largely favored, with moderate ties to moral conviction.
 
@@
 Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.
+The high emotional misinformation induces stronger emotions than the low emotion misinformation, especially for disgust.
+Both misinformation messages increase demand for restrictive policies, but only the high misinformation misinformation significantly increases moral conviction (by 6 percentage points).
+There is thus support for H1.
 
 ```{r misinfo-only, fig.align="center", fig.cap="Effects of misinformation with low and high emotional framing without following correction.
-Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="t", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 # H1: misinformation effect
@@
 
-```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="t", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 
@@
 
-Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe).
+Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe)).
 It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).
+On average, the two competitive frames seem to cancel each other out.
+This points to a challenge for science communication within an information environment in which accurate information (even if persuasive) has to compete with inaccurate or misleading information.
 
 ```{r counterframe,  fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing.
-Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
+Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="t", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
 

### 2021-09-22 data-preparation.R
load data from drive only if it is there

@@ -9,8 +9,9 @@ rm(list = ls())
 # load data (so far pre-test data)
 #load("Z:/3_current_Pretest/Pretest_W53.RData")
 
+if(exists(here("GIP_W53_V1.RData"))) {
 load(here("GIP_W53_V1.RData"))
-
+}
 # internal release
 if(!exists("GIP_W53_V1")) {
   load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")

### 2021-09-22 paper.Rmd
date

@@
   | GESS and SFB 884, University of Mannheim
-date: September 16, 2021
+date: \today
 linestretch: 1.5

### 2021-09-23 data-preparation.R
file.exist instead of exist

@@ -9,7 +9,7 @@ rm(list = ls())
 # load data (so far pre-test data)
 #load("Z:/3_current_Pretest/Pretest_W53.RData")
 
-if(exists(here("GIP_W53_V1.RData"))) {
+if(file.exists(here("GIP_W53_V1.RData"))) {
 load(here("GIP_W53_V1.RData"))
 }
 # internal release

### 2021-09-24 paper.Rmd
include moralization in the title

@@
 title: |
-  | Emotional framing and the effectiveness of 
-  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
+  | Emotional framing, moralization,
+  | and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study.
 This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on "The Political Economy of Reforms" (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]

### 2021-09-24 paper.Rmd
theory

@@
 
-*t.b.d.*
+Recent times have seen growing concerns about political misinformation in various forms such as "fake news" or the so called "infodemic" of science misinformation around the COVID-19 pandemic [@jerit2020political; @lazer2018science; @scheufele2021misinformation].
+Misinformed individuals hold misperceptions, i.e.
+"factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.
+128].
+Wide-spread belief in misinformation biases collective public opinion and thus poses a threat for democratic outcomes [@kuklinski2000misinformation].
+<!-- This concern is now more prevalent than ever.
+We live in a complex world and the digital era has made available an abundance of information at a glance.
+While this gives huge opportunities, concerns about "fake news" and its dissemination through digital technologies and social media are rampant [@allcott2017social; lazer2018science].
+-->
+Misperceptions can be due to both disinformation (false or misleading information disseminated with the intent to deceive) or misinformation (false information that may be disseminated with good intentions) [@southwell2018misinformation], as well as the flawed interpretation of accurate information, e.g.
+through cognitive biases [@lau2001advantages].
 
-* Definition of related terms: misinformation, disinformation, fake news, misperceptions [@flynn2017nature; @jerit2020political; @kuklinski2000misinformation].
-* Wide-spread misperceptions may distort public opinion and corrective information (fact-checking) is one way to tackle the problem.
-* People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
+<!-- * People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
+-->
+
+Because these misperceptions exist, both public and scholarly interest has been directed heavily to the question of how to correct them [@ec2018tackling; @lewandowsky2012misinformation; @walter2018unring}.
+Some studies find that directional motivation not only leads individuals to disregard corrective information, but produces a "backfire" effect where the incorrect prior belief is even strengthened despite having been corrected [@hart2012boomerang; @nyhan2010corrections].
+While being worrisome, this effect is rare.
+Predispositions such as ideology do affect the way corrective information is processed, but research also shows that corrections by and large work in the expected direction [@guess2018does; @wood2019elusive].
+A quickly growing number of studies accumulates evidence on the effects of corrective information and fact-checking in various contexts [@walter2018unring; @walter2019fact].
+
 * However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]

### 2021-09-24 paper.Rmd
limitation

@@
 Recent times have seen growing concerns about political misinformation in various forms such as "fake news" or the so called "infodemic" of science misinformation around the COVID-19 pandemic [@jerit2020political; @lazer2018science; @scheufele2021misinformation].
+Wide-spread belief in misinformation biases collective public opinion and thus poses a threat for democratic outcomes [@kuklinski2000misinformation].
 Misinformed individuals hold misperceptions, i.e.
@@
 128].
-Wide-spread belief in misinformation biases collective public opinion and thus poses a threat for democratic outcomes [@kuklinski2000misinformation].
 <!-- This concern is now more prevalent than ever.
@@
 -->
-Misperceptions can be due to both disinformation (false or misleading information disseminated with the intent to deceive) or misinformation (false information that may be disseminated with good intentions) [@southwell2018misinformation], as well as the flawed interpretation of accurate information, e.g.
-through cognitive biases [@lau2001advantages].
+While misinformation is often used as an umbrella term for inaccurate information, one can further specify disinformation as false or misleading information disseminated with the intent to deceive, whereas misinformation may be disseminated with good intentions [@southwell2018misinformation].
 
@@
 
-Because these misperceptions exist, both public and scholarly interest has been directed heavily to the question of how to correct them [@ec2018tackling; @lewandowsky2012misinformation; @walter2018unring}.
+As a response to concerns about various forms of misinformation both public and scholarly interest has been directed heavily to the question of how to correct them [@lewandowsky2012misinformation].
 Some studies find that directional motivation not only leads individuals to disregard corrective information, but produces a "backfire" effect where the incorrect prior belief is even strengthened despite having been corrected [@hart2012boomerang; @nyhan2010corrections].
 While being worrisome, this effect is rare.
-Predispositions such as ideology do affect the way corrective information is processed, but research also shows that corrections by and large work in the expected direction [@guess2018does; @wood2019elusive].
-A quickly growing number of studies accumulates evidence on the effects of corrective information and fact-checking in various contexts [@walter2018unring; @walter2019fact].
+Predispositions such as ideology do affect the way corrective information is processed, but research also shows that corrections work in the expected direction [@carnahan2021correcting; @guess2018does; @walter2019fact; @wood2019elusive].
 
-* However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
+However, while corrective information usually moves factual beliefs in the appropriate direction, other outcomes such as attitudes, blame attributions, or evaluations have been shown to be less receptive to corrections [@bisgaard2019getting; @nyhan2019taking; @swire2019they].
+
 * Especially troubling cases are "affective perseverance" [@sherman2002affective], and "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
@@
 ## Emotional framing and moralization {-}
-
-*t.b.d.*
 
@@
 \newpage
+
 # Conclusion
-
-*t.b.d.*
 
@@
 
-* No distinction between emotional framing and visual cues (see e.g. @hameleers2020picture).
+* In the experiment, I pictures as visual cues to convey emotional frames.
+While this manipulation seems to be working for the intended emotion of disgust, I cannot distinguish between the effect of emotional framing through images and the effect of visual cues per se.
+Research on the effects of visuals for dis- and corrective information shows that visual disinformation is only slightly more credible whereas visual correction is not more effective than textual correction [@hameleers2020picture].
+As an explanation, the authors specualte that the textual format works because it "fits fact-checkers' reliance on short, rational and eveidecne-based argumentation" [@hameleers2020picture, p.
+297].
+In their study, imapges are conceptualized as providing "indexality", i.e.
+a clearer link to reality, and not emotionality as in my study.
+Overall, the non-existence of an effect of visual cues as "indexality" on the effectiveness of correction provides support for my argument that in my study, it is the emotionality of the images and not the visual cues per se that make for the different effects, but further research is needed to completely disentangle these effects.
 

### 2021-09-24 paper.Rmd
mostly formatting

@@
 The information within the study can thus be classified as misinformation.^[This is consistent with a definition of resulting mis*perceptions* as "factual beliefs that are false or contradict the best available evidence in the public domain" [@flynn2017nature, p.128].
-If one would presume the intent to mislead, this would then be classified as disinformation.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.
+It should be noted, however, that the treatment material in @clifford2019emotional uses this example only as one of several and remains vague without giving specific scientific information.] I vary this misinformation to be either low or high in emotional framing by ex- or including graphical images of a rat with a cancer tumor and rotten corn, images that are taken from @clifford2019emotional and shown to induce negative feelings, specifically disgust.
 
@@
 
-Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects.
-On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint.
-Anxiety was above the midpoint with a mean of 2.4 whereas disgust was relatively low with a mean of 1.3, and sadness again moderately below the midpoint with 1.7.
-These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.
-
-With regard to the factual belief that GM food is safe to eat, the sample is skewed toward the view that this is rather not true (mean = 3.2 on a scale from 1 to 7).
-When it comes to opinions about policy restrictions for GM food, respondents are rather in favor (mean = 5.2 on a scale from 1 to 7).
-Finally, the measure of moral conviction is close to the midpoint (mean = 4.4 on a scale from 0 to 8).
-These results indicate that the misperception that GM food is not safe for one own's health is wide-spread in Germany and that policy restrictions are largely favored, with moderate ties to moral conviction.
-
 ```{r descriptives, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
+Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects.
+On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint.
+Anxiety was above the midpoint with a mean of 2.4 whereas disgust was relatively low with a mean of 1.3, and sadness again moderately below the midpoint with 1.7.
+These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.
+
+With regard to the factual belief that GM food is safe to eat, the sample is skewed toward the view that this is rather not true (mean = 3.2 on a scale from 1 to 7).
+When it comes to opinions about policy restrictions for GM food, respondents are rather in favor (mean = 5.2 on a scale from 1 to 7).
+Finally, the measure of moral conviction is close to the midpoint (mean = 4.4 on a scale from 0 to 8).
+These results indicate that the misperception that GM food is not safe for one own's health is wide-spread in Germany and that policy restrictions are largely favored, with moderate ties to moral conviction.
+
 ## Moralization through misinformation {-}
-
-I now turn to the effects of the informational treatments on the outcome variables.
-To quantify these effects, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.
-For a more intuitive interpretation of the effect sizes - and to be able to compare outcome variables measured on different scales - I rescale all outcome variables to range from 0 to 1.
-I do not include any control variables because the random assignment of the experimental design provides the better safeguard against any possible confounders.
-
-In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.
-Without any controls, the coefficients refer to the difference in the mean of the outcome between the control and the respective treatment group.
-^[The significance of the coefficients is equivalent to running separate t-tests between the group means, which raises the issue of adjusting for multiple testing.
-There are different approaches from different traditions to that.
-When conducting many independent t-tests, it seems straight-forward to adjust for multiple testing, although the general advise to always adjust for multiple testing is not warranted [@althouse2016adjust].
-Another approach is to conduct ANOVA, which is an omnibus test, with the null-hypothesis of no effect of the treatment conditions together versus the control.
-To check for individual differences between experimental groups, post-hoc tests can be conducted with respective appreciation of multiple testing adjustments.
-This is appropriate if one has no specific hypotheses to begin with.
-If one has these hypotheses, however, one can use ANOVA with planned contrasts (e.g.
-[@field2012discovering]), which is then again equivalent to the results of a regression with dummy variables coded according to the specified contrasts.
-When I use multiple outcomes in my pre-registered hypotheses, the expectation is not that any of the outcomes shows an effect (in which case adjustment for multiple testing should be done), but rather a joint pattern of effects.]
 
@@
 
-First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message.
-The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message.
-The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
+I now turn to the effects of the informational treatments on the outcome variables.
+To quantify these effects, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.
+For a more intuitive interpretation of the effect sizes - and to be able to compare outcome variables measured on different scales - I rescale all outcome variables to range from 0 to 1.
+I do not include any control variables because the random assignment of the experimental design provides the better safeguard against any possible confounders.
 
-In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome.
-Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group.
-For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.
-
-Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.
-The high emotional misinformation induces stronger emotions than the low emotion misinformation, especially for disgust.
-Both misinformation messages increase demand for restrictive policies, but only the high misinformation misinformation significantly increases moral conviction (by 6 percentage points).
-There is thus support for H1.
+In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.
+Without any controls, the coefficients refer to the difference in the mean of the outcome between the control and the respective treatment group.
+^[The significance of the coefficients is equivalent to running separate t-tests between the group means, which raises the issue of adjusting for multiple testing.
+There are different approaches from different traditions to that.
+When conducting many independent t-tests, it seems straight-forward to adjust for multiple testing, although the general advise to always adjust for multiple testing is not warranted [@althouse2016adjust].
+Another approach is to conduct ANOVA, which is an omnibus test, with the null-hypothesis of no effect of the treatment conditions together versus the control.
+To check for individual differences between experimental groups, post-hoc tests can be conducted with respective appreciation of multiple testing adjustments.
+This is appropriate if one has no specific hypotheses to begin with.
+If one has these hypotheses, however, one can use ANOVA with planned contrasts (e.g.
+[@field2012discovering]), which is then again equivalent to the results of a regression with dummy variables coded according to the specified contrasts.
+When I use multiple outcomes in my pre-registered hypotheses, the expectation is not that any of the outcomes shows an effect (in which case adjustment for multiple testing should be done), but rather a joint pattern of effects.]
 
@@
 
+First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message.
+The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message.
+The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.
+
+In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome.
+Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group.
+For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.
+
+Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.
+The high emotional misinformation induces stronger emotions than the low emotion misinformation, especially for disgust.
+Both misinformation messages increase demand for restrictive policies, but only the high misinformation misinformation significantly increases moral conviction (by 6 percentage points).
+There is thus support for H1.
+
 ## Effectiveness of neutral correction {-}
-
-After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
-The neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation.
-The emotional frame of the misinformation therefore moderates the correcion effectiveness.
-Although weakened, the effects on policy opinion and moral conviction induced by the high emotion misinformation persist after the correction, confirming H2b.
-Even if the correction significantly moves the factual belief after the high emotion misinformation, there still persists an effect of the misinformation relative to the control group.
-The correction is therefore not successful in eradicating the misinformation effect, which rejects H2a.
-In sum, there is no belief echo, but a moderation of the correction effectiveness.
 
@@
 
+After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
+The neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation.
+The emotional frame of the misinformation therefore moderates the correcion effectiveness.
+Although weakened, the effects on policy opinion and moral conviction induced by the high emotion misinformation persist after the correction, confirming H2b.
+Even if the correction significantly moves the factual belief after the high emotion misinformation, there still persists an effect of the misinformation relative to the control group.
+The correction is therefore not successful in eradicating the misinformation effect, which rejects H2a.
+In sum, there is no belief echo, but a moderation of the correction effectiveness.
+
 ## Counterframe correction {-}
-
-Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe)).
-It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).
-On average, the two competitive frames seem to cancel each other out.
-This points to a challenge for science communication within an information environment in which accurate information (even if persuasive) has to compete with inaccurate or misleading information.
 
@@
 
-\newpage
+Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe)).
+It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).
+On average, the two competitive frames seem to cancel each other out.
+This points to a challenge for science communication within an information environment in which accurate information (even if persuasive) has to compete with inaccurate or misleading information.
 
@@
 
-* In the experiment, I pictures as visual cues to convey emotional frames.
+* In the experiment, I use pictures as visual cues to convey emotional frames.
 While this manipulation seems to be working for the intended emotion of disgust, I cannot distinguish between the effect of emotional framing through images and the effect of visual cues per se.
 Research on the effects of visuals for dis- and corrective information shows that visual disinformation is only slightly more credible whereas visual correction is not more effective than textual correction [@hameleers2020picture].
-As an explanation, the authors specualte that the textual format works because it "fits fact-checkers' reliance on short, rational and eveidecne-based argumentation" [@hameleers2020picture, p.
+As an explanation, the authors specualte that the textual format works because it "fits fact-checkers' reliance on short, rational and evidence-based argumentation" [@hameleers2020picture, p.
 297].
-In their study, imapges are conceptualized as providing "indexality", i.e.
+In their study, imapges are conceptualized as providing "indexicality", i.e.
 a clearer link to reality, and not emotionality as in my study.
-Overall, the non-existence of an effect of visual cues as "indexality" on the effectiveness of correction provides support for my argument that in my study, it is the emotionality of the images and not the visual cues per se that make for the different effects, but further research is needed to completely disentangle these effects.
+Overall, the non-existence of an effect of visual cues as "indexicality" on the effectiveness of correction provides support for my argument that in my study, it is the emotionality of the images and not the visual cues per se that make for the different effects, but further research is needed to completely disentangle these effects.
 

### 2021-09-28 data-preparation.R
include demographics from core questionnaire

@@ -4,24 +4,31 @@ library(sjmisc)
 library(scales)
 library(here)
 
+# make sure to start with a clean workspace
 rm(list = ls())
 
-# load data (so far pre-test data)
-#load("Z:/3_current_Pretest/Pretest_W53.RData")
+# load data 
+# so far, internal release on the sfb drive
+# later (in december) data will be made public through the GESIS data archive
+# https://dbk.gesis.org/dbksearch/GDesc2.asp?no=0109&tab=&ll=10&notabs=1&db=E
 
+# if data is already on hard drive, load into workspace
 if(file.exists(here("GIP_W53_V1.RData"))) {
 load(here("GIP_W53_V1.RData"))
 }
-# internal release
+
+# if data is not in the workspace, load from internal data release
 if(!exists("GIP_W53_V1")) {
   load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
 }
 
-# later, the data will be made public
-
+# put into data object for more intuitive handling while leacing the raw data intact
 data <- GIP_W53_V1
 names(data)
 
+############################
+# recode variables
+############################
 
 # pre-attitude
 frq(data$CE53482)
@@ -182,4 +189,81 @@ frq(data$info_search)
 # select relevant data
 data <- subset(data, select= c(id_g, pre_attitude:info_search))
 
-#save(data, file = "./data/pre-test.R")
+############################
+# add demographics from core questionnaire
+############################
+
+# if data is already on hard drive, load into workspace
+if(file.exists(here("GIP_W49_V21.RData"))) {
+  load(here("GIP_W49_V2.RData"))
+}
+
+# if data is not in the workspace, load from internal data release
+if(!exists("GIP_W49_V21")) {
+  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W49_V2.Rdata")
+}
+
+names(GIP_W49_V2)
+
+core <- GIP_W49_V2
+
+# gender
+frq(GIP_W49_V2$gender_20)
+core$gender <- fct_collapse(core$gender_20,
+                            "male" = "1. männlich",
+                            "female" = "2. weiblich",
+                            NULL = c("-97. trifft nicht zu",
+                                     "-91. 'bitte wählen'-Platzhalter",
+                                     "-90. item nonresponse",
[…]

### 2021-09-28 paper.Rmd
name chunks

@@
 
-```{r, echo = FALSE}
+```{r main-analysis, echo = FALSE}
 # linear regression models with 0-1 rescaled DV

### 2021-09-30 paper.Rmd
some possible to dos

@@
 ```
+
+\newpage
+# To Do
+
+- To what audience should the paper be framed / be the focus of the theoretical discussion?
+(Probably misinformation and corrective information, but it may also be necessary to engage a bit more with literature on moralization of issues, and emotions and information processing)
+- Closer link of theory and empirical expecations, e.g.
+from affective intelligence theory or cognitive appraisal approaches.
+There probably is an interesting tension here: emotion of disgust should lead to less engagement with following information because of a general averse reaction of "get away from me", whereas emotion of anxiety should lead to a closer processing of incoming information.
+In real life, anxiety and disgust may often go together, as happens empirically in the experiment.
+- Analysis: what about treatment heterogeneity? There is a pre-attitude item here to exploit, and also possible other variables from previous GIP panel waves
+- Analysis: mabe show the moderation effect of the emotional frame directly
+- For the discussion and further avenues: What role do elite cues play in all of that?
+So, e.g., political elits may work with strong emotion in framing disinformation, but the scientific / fact-checking elites do not use strong emotions in their communication (is that true empirically?)
+- Exploratory: there are also ambivalence items and an information search item, which are not yet part of the paper
 

### 2022-07-15 paper.Rmd
Update paper.Rmd

@@
 ```
-
+```
+![p_h1](https://user-images.githubusercontent.com/49069412/179274284-155b24ef-33e3-4d2c-bc6d-2c9d16fba174.png)
+```
 First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message.

### 2022-07-22 paper.Rmd
Added pathways
Commented out pathways to data, now works smoothly locally with paper.Rmd

--- 2021-09-30 paper.Rmd
@@
 
-#sjPlot::save_plot("./figures/p_h1.png", width = 24, height = 16)
+sjPlot::save_plot("./p_h1.png", width = 24, height = 16)
+```
 
-```
+![p_h1](./p_h1.png)
 
@@
 
+sjPlot::save_plot("./p_neutral_correction.png", width = 24, height = 16)
+
 #sjPlot::save_plot("./figures/p_neutral_correction.png", width = 24, height = 16)
@@
 
+![p_neutral_correction](p_neutral_correction.png)
 After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).
@@
 
-#sjPlot::save_plot("./figures/p_counter_frame.png", width = 24, height = 16)
+sjPlot::save_plot("./p_counter_frame.png", width = 24, height = 16)
 
 ```
+
+![p_counter_frame1](./p_counter_frame.png)
 
@@
 
-#sjPlot::save_plot("./figures/p_info_effect.png", width = 24, height = 9)
+sjPlot::save_plot("./p_info_effect.png", width = 24, height = 9)
 
 ```
-
+![p_info_effect](./p_info_effect.png)
 \newpage
@@
 ```{r screenshot-low-misinfo, fig.align = 'center', out.width = "100%", fig.cap = "Low emotion misinformation", echo = FALSE}
-knitr::include_graphics("./figures/misinfo_low_emotion.png")
+knitr::include_graphics("./misinfo_low_emotion.png")
 ```
@@
 ```{r screenshot-neutral-correction, fig.align = 'center', out.width = "100%", fig.cap = "Neutral correction", echo = FALSE}
-knitr::include_graphics("./figures/neutral_correction.png")
+knitr::include_graphics("./neutral_correction.png")
 ```

### 2022-07-22 data-preparation.R

@@ -13,13 +13,18 @@ rm(list = ls())
 # https://dbk.gesis.org/dbksearch/GDesc2.asp?no=0109&tab=&ll=10&notabs=1&db=E
 
 # if data is already on hard drive, load into workspace
-if(file.exists(here("GIP_W53_V1.RData"))) {
-load(here("GIP_W53_V1.RData"))
-}
+#if(file.exists(here("GIP_W53_V1.RData"))) {
+#load(here("GIP_W53_V1.RData"))
+#}
 
 # if data is not in the workspace, load from internal data release
+#if(!exists("GIP_W53_V1")) {
+#  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+#}
+
 if(!exists("GIP_W53_V1")) {
-  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+  load("/Volumes/data$/2_Data/data_for_R_users/GIP_W53_V1.Rdata")
+  
 }
 
 # put into data object for more intuitive handling while leacing the raw data intact
@@ -194,15 +199,18 @@ data <- subset(data, select= c(id_g, pre_attitude:info_search))
 ############################
 
 # if data is already on hard drive, load into workspace
-if(file.exists(here("GIP_W49_V21.RData"))) {
-  load(here("GIP_W49_V2.RData"))
-}
+#if(file.exists(here("GIP_W49_V21.RData"))) {
+#  load(here("GIP_W49_V2.RData"))
+#}
 
 # if data is not in the workspace, load from internal data release
+#if(!exists("GIP_W49_V21")) {
+#  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W49_V2.Rdata")
+#}
 if(!exists("GIP_W49_V21")) {
-  load("//sfb884-share.ad.uni-mannheim.de/data$/2_Data/data_for_R_users/GIP_W49_V2.Rdata")
+  load("/Volumes/data$/2_Data/data_for_R_users/GIP_W49_V2.Rdata")
+  
 }
-
 names(GIP_W49_V2)
 
 core <- GIP_W49_V2
